\mainmatter


# Event History and Survival Data

```{r settings1, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(comment = "", message = FALSE, echo = TRUE, cache = FALSE)
options(width = 70, digits = 7)
```


The thing that characterizes event history and survival data the most
is its *dynamic* nature. Individuals are
followed over
time, and during that course, the timing of events of interest are
noted. Naturally, things may happen that makes it necessary to interrupt an
individual follow-up, such as the individual suddenly disappearing for some
 reason. With classical statistical tools, such as
linear regression, these observations are difficult or impossible to handle
in the analysis. The methods discussed in this book aim among other
things, at solving such problems.


In this introductory chapter, the use of the special techniques that
constitute survival and event history analysis are motivated. The concepts of
*right censoring*\index{censoring!right} and
*left truncation*\index{left truncation} are defined and discussed.
The data sets used throughout the book are also presented
in this chapter.

The environment R is freely (under a *GPL license*) available for
download from [CRAN](https://cran.r-project.org). There you will find
precompiled versions for Linux, MacOS X, and Microsoft Windows, as well as
the full source code, which is open. See *Appendix C* for more
information.

## Survival data

Survival data (survival times)\index{survival data} constitute the
simplest form of event history data\index{event history data}. A survival
time is defined as the
time it takes for an event to occur, measured from a well-defined start
event. Thus, there are three basic elements which must be well defined: a
time origin, a scale for measuring time, and an event. The response in a
statistical analysis of such data is the exact time elapsed from the time
origin to the time at which the event occurs. The challenge, which
motivates special methods, is that in most applications, this
duration is often not possible to observe exactly.

As an introduction to the research questions that are suitable for handling
with event history and survival analysis, let us look at a data set found
in the **eha**
package [@eha] in R [@cran].

```{example, label = "oldmort01", name = "Old age mortality", echo = TRUE}
```
\index{mortality!old age|(}
The data set *oldmort*\index{Data sets!old age mortality|(} in **eha**
contains survival data from the
parish Sundsvall\index{Sundsvall} in the mid-east of 19th century
Sweden. The name *oldmort* is an acronym for *old age
  mortality*. The source is digitized information from historical parish registers, church books. More
information about this can be found at the web page of the
*Centre for Demographic and Ageing Research* at Umeå University
([CEDAR](http://www.cedar.umu.se)).

The sampling was done as follows: Every person who was present and alive
and 60 years of age or above anytime between 1 January 1860 and 31 December
1879 was followed from the entrance age (for most people that would be 60)
until the age when last seen, determined by death, out-migration, or
surviving until 31 December 1879. Those born during the
eighteenth century would enter observation at an age above 60, given that
they lived long enough, that is at least until January 1, 1860.

Two types of finishing the observation of a person are distinguished:
Either it is by
*death* or it is by something else, out-migration or end of study period.
In the first case we say that the event of interest has occurred, in the
second case not.



After installing the **eha** package and starting an R session (see
Appendix C), the data set is loaded by loading **eha** as follows.
```{r oldmort1, comment = ""}
library(eha)
```
Let us look at the first few lines of *oldmort* (slightly manipulated). 
It is conveniently done with the aid of the R function **head**:

```{r manip1, echo=FALSE}
oldmort$id <- as.integer(as.factor(oldmort$id))
##oldmort$birthdate <- toDate(oldmort$birthdate)
```

```{r head1, echo = TRUE}
head(oldmort, 3)
```

Note that the first printed column, without a name, contains the *row names*. If
no row names are explicitly given, they are set to the *row numbers*. If you don't
want row names printed, you have to use the more cumbersome 

```{r headsubst1, echo = TRUE}
print(oldmort[1:3, ], row.names = FALSE)
```

There are several packages for nice printing of tables, if you want that.

The variables in *oldmort* have the following definitions and
interpretations:

*   **id** A unique id number for each individual.
*   **enter, exit** The start age and stop age for this record
  (spell). For instance, in row No.\ 1, individual No.
  `r oldmort$id[1]` enters under observation at age
  `r round(oldmort$enter[1], 2)` and exits at age `r round(oldmort$exit[1], 2)`. Age is
  calculated as the number of days elapsed since birth and this number is
  then divided by 365.25 to get age in years. The denominator is the
  average length of a year, taking into account that every fourth year is
  366 days long.
  The first individual was born on June 27, 1765, and so almost 95 years
  of age when the study started. Suppose that this woman had died at age
  94; then she had not been in our study at all. This property of our
  sampling procedure introduces a phenomenon known as
  *length-biased sampling*\index{length-biased sampling}. That is,
  of those born in the eighteenth century, only those who live well beyond
  60 will be included. This bias must be compensated for in the analysis,
  and it is accomplished by conditioning on the fact that these
  persons were alive at January 1, 1860. This technique is called
  *left truncation*\index{left truncation}.
*   **event** A logical variable (taking values *TRUE* or *FALSE*)
indicating if the exit is a death (*TRUE*) or not (*FALSE*). For our first
individual, the value is *TRUE*, indicating
  that she died at the age of 95.81 years. Another possible coding is to use
  *event = 1* for a death and *event = 0* otherwise.
*   **birthdate** The birth date expressed as the time (in years) elapsed
  since January 1, year 0 (which by the way does not exist). For instance,
  the (pseudo) date 1765.490 is really June 27, 1765. The fraction 0.490 is
  the fraction of the year 1765 that elapsed until the birth of individual
  No. `r oldmort$id[1]`. However, here it is printed as a real date.
*   **m.id** *Mother's id*. It is unknown for all the individuals listed
  above. That is the symbol *NA*\index{NA}, which stands for *Not Available*.
  The oldest people in the data set typically have no links
  to parents.
*   **f.id** *Father's id*. See *m.id*.
*   **sex** A categorical variable\index{categorical variable}  with
  the levels *female* and *male*.
*   **civ** Civil status. A categorical variable\index{categorical variable|(}
  with three levels;
  *unmarried*, *married*, and *widow(er)*.
*   **ses.50** Socio-economic status (SES) at age 50. Based on occupation
  information. There is a large proportion of *NA* (missing
  values\index{missing values}) in this variable. This
  is quite natural, because this variable was of secondary interest to the
  record holder (the priest in the parish). The occupation is only noted in
  connection to a vital event in the family (such as a death, birth,
  marriage, or in- or out-migration). For those who were above 50 at the start
  of the period there is no information on SES at 50.
*   **birthplace** A categorical variable with two categories, *parish*
  and *remote*, representing born in parish and born
    outside parish, respectively.
*   **imr.birth** A rather specific variable. It measures the infant
    mortality rate in the birth parish at the time of birth (per cent).
*   **region** Present geographical area of residence. The parishes in the
  region are grouped into three regions, *Sundsvall town*, *rural*,
  and *industry*. The industry is the sawmill one, which grew rapidly
  in this
  area during the late part of the 19th century. The Sundsvall area was in
  fact one of the largest sawmill areas in Europe at this time.
  \index{categorical variable|)}

Of special interest is the triple `(enter, exit, event)`, because it
represents the *response variable*, or what can be seen of it. More
specifically, the sampling frame is all persons observed to be alive and
above 60 years of age between 1 January 1860 and 31 December 1879. The
start event for these individuals is their 60th anniversary and the stop
event is death. Clearly, many individuals in the data set did not die
before 1 January 1880, so for them we do not know the full duration
between the start and stop events; such individuals are said to be
*right censored* (the exact meaning of which will be given soon). The
third component in the survival object `(enter, exit, event)`,
 `event`, is a *logical* variable taking the value `TRUE`
if `exit` is the true duration (the interval ends with a death) and
`FALSE` if the individual is still alive at the duration "last seen".

Individuals aged 60 or above between 1 January 1860 and 31 December 1879
are included in the
study. Those who are above 60 at this start date are included only if they
did not die between
the age of 60 and the age at 1 January 1860. If this is not taken into
account, a bias in the estimation of mortality will result. The proper way
of dealing with this problem is to use *left truncation*, which is
indicated by the variable `enter`. If we look at the first rows of *oldmort* we
see that the `enter` variable is very large; it is the
age for each individual at 1 January 1860. You can add `enter` and *birthdate*
for the first six individuals to see that:
```{r add1}
oldmort$enter[1:6] + oldmort$birthdate[1:6]
```

The statistical implication (description) of left truncation is that its
presence forces the analysis to be
*conditional*\index{conditional probability} on survival up to the
age `enter`.

A final important note: In order to get the actual duration at exit, we
must subtract 60 from the value of `exit`. When we actually perform a
survival analysis in R, we should subtract 60 from both `enter` and
`exit` before we begin. It is not absolutely necessary in the case of Cox
regression, because of the flexibility of the baseline hazard in the model
(it is in fact left unspecified!). However, for parametric models, it may
be important in order to avoid dealing with truncated *distributions*.

Now let us think of the research questions that could be answered by
analyzing this data set. Since the data contain individual information on
the length of life after 60, it is quite natural to study what determines a
long life and what are the conditions that are negatively correlated with
long life. Obvious questions are: (i) Do women live longer than men? (Yes),
(ii) Is it advantageous for a long life to be married? (Yes), (iii) Does
socio-economic status play any role for a long life? (Don't know), and (iv)
Does place of birth have any impact on a long life, and if so, is it
different for women and men?$\Box$

The answers to these, and other, questions will be given later. The methods
in later chapters of the book are all illustrated on a few core
examples. They are all presented a first time in this chapter.
\index{Data sets!old age mortality|)}

The data set *oldmort* contained
only two states, referred to as *Alive* and *Dead*, and
one possible transition, from Alive to Dead, see Figure&nbsp;\@ref(fig:survival).

```{r survival, fig.cap = "Survival data.", echo = FALSE,fig.height=4}
##knitr::include_graphics("images/fig1.1.pdf")
##source("R/fig1.1.R")
##fig1.1()
plot(1, 1, type = "n", xlim = c(0, 20), ylim = c(0, 10),
   axes = FALSE,
   xlab = "", ylab = "")
symbols(c(5, 15), c(5, 5), circles = c(2, 2), inches = FALSE, add = TRUE)
arrows(7, 5, x1 = 13, length = 0.1)
text(5, 5, "Alive")
text(15, 5, "Dead")
```

The ultimate study object
in survival analysis is the time it takes from entering state *Alive*
(e.g., becoming 60 years of age) until entering state *dead* (e.g.,
death). This time
interval is defined by the exact time of two events, which we may call
*birth* and *death*, although in practice these two events may be
almost any kind of events. Economists, for instance, are interested in the
duration of out-of-work spells, where "birth" refers to the event of losing
the job, and "death" refers to the event of getting a job. In a clinical
trial regarding treatment of cancer, the starting event time may be time of
operation, and the final event time is time of relapse (if any).

## Right censoring

When an individual is lost to follow-up, we say that she is *right censored*
\index{censoring!right}, see Figure&nbsp;\@ref(fig:rightcensoring).

```{r rightcensoring, fig.cap = "Right censoring", echo = FALSE, fig.height= 3}
plot(c(0, 10), c(2, 2), type = "l", ylim = c(0, 4),
     xlab = "Duration", xlim = c(0, 20), ylab = "", axes = FALSE)
axis(1, at = c(0, 6, 10), labels = c("0", "C", "T"))
box()
rect(6, 1, 28, 3, col = "grey")
lines(c(0, 10), c(2, 2), cex = 1.5)
text(13, 2, "+      Censored")
abline(v = 0)
```


As indicated in Figure&nbsp;\@ref(fig:rightcensoring), the true age at death is
*T*, but due to right
censoring, the only information available is that death age *T* is larger
than *C*. The number *C* is the age at which this individual was last
seen. In ordinary, classical regression analysis, such data are
difficult, if not impossible, to handle. Discarding such information may
introduce bias. The modern theory of survival analysis offers simple ways
to deal with right censored data.

A natural question to ask is: If there is right censoring, there should be
something called *left* censoring, and if so, what is it? The answer
to that is that yes, *left censoring*\index{censoring!left} refers to
a situation where the
only thing known about a death age is that it is *less than* a certain
value $C$. Note carefully that this is different from *left
  truncation*, see the next section.

## Left truncation

The concept of *left truncation*\index{left truncation|(}, or
*delayed entry*, is well illustrated by the data set *oldmort*
that was discussed in detail in Example \@ref(exm:oldmort01).
Please note the
difference compared to *left censoring*\index{censoring!left}. 
Unfortunately, you may still see articles where these two concepts are mixed up.

It is illustrative to think of the construction of the data set
  *oldmort* as a statistical follow-up study, starting on January 1,
1860. At that day, all persons present in the parish and 60 years
of age *or above*, are included in the study. It is decided that the
study will end at 31 December 1879, that is, the study period (follow-up
time) is 20 years. The interesting event in this study is
*death*. This means that the start event is the sixtieth anniversary
of birth and the final event is death. Due to the calendar time constraints
(and migration), all individuals will not be observed to die (especially
those who live long), and moreover, some individuals will enter the study
after the "starting" event, the sixtieth anniversary. A person who enter
late, say he is
65 on January 1, 1860, had not been included had he died at age 63
(say). Therefore, in the analysis, we must
*condition*\index{conditional probability} on the fact
that he was alive at 65. Another way of saying this is to say that this
observation is *left truncated* at age 65.

People being too young at the start date will be included from the day they
reach 60, if that happens before the closing date, December 31, 1879. They
are not left truncated, but will have a higher and higher probability of
being right censored, the later they enter the study.
\index{left truncation|)}

## Time scales

\index{time scale|(}

In demographic applications *age* is often a natural time
scale\index{time scale}, that
is, time is measured from birth. In the old age data just discussed, time
was measured from age 60 instead. In a case like this, where there is a
common ``late start age'', it doesn't matter much, but in other situations
it does. Imagine for instance that interest lies in studying the time it
takes for a woman to give birth to her first child after marriage. The
natural way of measuring time is to start the clock at the day of marriage,
but a possible (but not necessarily recommended!) alternative is to start
the clock at some (small) common age of the women, for instance at
birth. This would give left truncated (at marriage)
observations, since women were sampled at marriage. There are two clocks
ticking, and you have to make a choice.

Generally, it is important to realize that there often are alternatives,
and that the
result of an analysis may depend strongly on the choice made.
\index{time scale|)}


### The Lexis diagram

\index{Lexis diagram|(}
Two time scales are nearly always present in demographic research: Age (or
duration) and calendar time. For instance, an investigation of mortality
may be limited in these two directions. In Figure \ref{fig:lexis1} this is
illustrated for a study of old age mortality during the years 1829 and
1895. ``Old age mortality'' is defined as mortality from age 50 and onwards
to age 100.
The Lexis diagram is a way of showing the interplay between the two time
scales and (human) life lines. Age moves vertically and calendar time
horizontally, which will imply that individual lives will move
diagonally, from birth to death, from south-west to north-east, in the
Lexis diagram. In our example study, we are only interested in the part of
the life lines that appear inside the small rectangle.

```{r lexis1,fig.cap = "Lexis diagram: time period 1829--1894 and ages 50--100.",fig.scap = "A lexis diagram", echo=FALSE,height=5.4}
source("R/lexis.R")
nul <- lexis()
```

Assume that the data set at hand is saved in the text file 'lex.dat'. Note
that this data set is not part of **eha**; it is only used here for
the illustration of the Lexis diagram.

```{r read.lex}
lex <- read.table("Data/lex.dat", header = TRUE)
lex
```
How do we restrict the data to fit into the rectangle given by the Lexis
diagram in Figure&nbsp;\@ref(fig:lexis1)? With the two functions **age.window**
and **cal.window** it is easy. The former fixes the 'age cut' while the
latter makes the 'calendar time cut'.

The age cut:
```{r age.cut}
lex <- age.window(lex, c(50, 100))
lex
```
Note that individual No.\ 6 dropped out completely because she died too young.
Then the calendar time cut:
```{r cal.cut}
lex <- cal.window(lex, c(1829, 1895))
lex
```
and here individual No.\ 7 disappeared because she died before January 1,
1829. Her death date is her birth date plus her age at death, $1740.513 +
74.112 = 1814.625$, or August 17, 1814.
\index{Lexis diagram|)}

## Event history data

Event history data arise, as the name suggests, by following
subjects over time and making notes about what happens and when. Usually
the interest is concentrated to a few specific kinds of events. The main
application in this book is demography and epidemiology, hence events of
primary interest
are *births*, *deaths*, *marriages* and *migration*.


```{example, name = "Marital fertility in 19th century Sweden", label = "marfert", echo = TRUE}
```
As a rather complex example, let us look at *marital fertility* in
19th century Sweden, see
Figure \@ref(fig:fert).

```{r fert, fig.cap = "Marital fertility.", echo = FALSE}
plot(1, 1, ylim = c(0, 20), xlim = c(0, 30), axes = FALSE,
     xlab = "", ylab = "", type = "n")

rect(1, 1, 8, 5)
text(4.5, 3, "4 children")
rect(1, 8, 8, 12)
text(4.5, 10, "5+ children")
rect(1, 15, 8+5, 19)
text(7, 17, "Married, no children")
##
add <- 20
rect(1 + add, 1, 8 + add, 5)
text(4.5 + add, 3, "3 children")
rect(1 + add, 8, 8 + add, 12)
text(4.5 + add, 10, "2 children")
rect(1 + add, 15, 8 + add, 19)
text(4.5 + add, 17, "1 child")
symbols(14.5, 10, circles = 2, add = TRUE, inches = FALSE)
text(14.5, 10, "'End'")
arrows(x0 = 13, y0 = 17, x1 = 21, length = 0.1)
arrows(x0 = 25, y0 = 15, y1 = 12, length = 0.1)
arrows(x0 = 25, y0 = 8, y1 = 5, length = 0.1)
arrows(x0 = 21, y0 = 3, x1 = 8, length = 0.1)
arrows(x0 = 4.5, y0 = 5, y1 = 8, length = 0.1)
arrows(x0 = 13, y0 = 15, x1 = 13.8, y1 = 12.4, length = 0.1)
arrows(x0 = 21, y0 = 15, x1 = 15.7, y1 = 12.0, length = 0.1)
arrows(x0 = 21, y0 = 10, x1 = 16.5, length = 0.1)
arrows(x0 = 21, y0 = 5, x1 = 15.7, y1 = 8, length = 0.1)
arrows(x0 = 8, y0 = 5, x1 = 13.2, y1 = 8, length = 0.1)
```

In a marital fertility study, women are typically followed over time from
the time of their marriage until the time the marriage is dissolved or her
fertility period is over, say at age 50, whichever comes first. The
marriage dissolution may be due to the death of the woman or of her
husband, or it may be due to a divorce. If the study is limited to a given
geographical area, women may get lost to follow-up due to
out-migration. This event gives rise to a
*right-censored*\index{censoring!right} observation.

During the follow-up, the exact timings of child births are
recorded. Interest in the analysis may lie in investigating which factors,
if any, that affect the length of birth intervals. A data set may look like
this:
```{r bi,echo=FALSE}
#load("../../bi.RData")
#bi[1:7, ]
require(eha)
data(fert)
n.rows <- 9
print(fert[1:n.rows, -9], row.names = FALSE)
```
This is the first `r n.rows` rows, corresponding to the first two
mothers in the
data file.
The variable
*id* is *mother's id*, a label that uniquely identifies each
individual.

A birth interval has a start point (in time) and an end point. These points
are the time points of births, except for the first interval, where
the start point is time of marriage, and the last interval, which is
open to the right. However, the last interval is stopped at the time of
marriage dissolution
or when the mother becomes 50, whatever comes first. The variable
  *parity* is zero for the first interval, between date of marriage and date
of first birth, one for the next interval, and so forth. The last (highest)
number is thus equal to the total number of births for a woman during her
first marriage (disregarding twin births, etc.).

Here is a description variable by variable of the data set.

*   **id** The mother's unique id.
*   **parity** Order of *previous* birth, see above for details. Starts
  at zero.
*   **age** Mother's age at the event defining the start of the interval.
*   **year** Calendar year for the birth defining the start of the interval.
*   **next.ivl** measures the time in years from the birth at *parity*
to the birth at *parity + 1*, or, for the woman's last interval, to the
age of right censoring.
*   **event** is an indicator
for the interval ending with a birth. It is always equal to 1, except for
the last interval, which always has event equal to zero.
*   **prev.ivl** is the length of the interval preceding this one. For the
  first interval of a woman, it is always *NA* (Not Available).
*   **ses** Socio-economic status (based on occupation data).


Just to make it clear: The first woman has id 1. She is
represented by two records, meaning that she gave birth to one
child. She waited 0.411 years from marriage to the first birth, and 22.348
years from the first birth to the second, *which never happened*.
 The second woman (2) is represented by
seven records, implying that she gave birth to six children. And so on.

Of course, in an analysis of birth intervals we are interested in
causal\index{causality}
effects; why are some intervals short while others are long?  The
dependence of the
history can be modeled by *lengths of previous intervals* (for the same
mother), *parity*, *survival of earlier births*, and so on. Note that all
relevant covariate information *must refer to the past*. More about
that later.


The first interval of a woman is different from the others, since it starts
with marriage. It therefore makes sense to analyze these intervals
separately. The last interval of a woman is also special; it always ends
with a right censoring, at the latest when the woman is 50 years of
age. You should think of data for a woman generated sequentially in time,
starting at the day of her marriage. Follow-up is made to the next birth,
as long as she is alive, the marriage is still alive, and she is younger
than 50 years of age. If there is no next birth, i.e., she reaches 50, or
the marriage is dissolved (most often by death of one of the spouses), the
interval is censored at the duration when she still was under
observation. Censoring can also occur by emigration, and reaching the end
of follow-up, in this case November 5, 1901.$\Box$

```{example, name = "The illness-death model", label = "illdeath"}
```
\index{illness-death model|(}
Another useful setup is the *illness-death* model, see
Figure \@ref(fig:illdeathfig).


```{r illdeathfig, fig.cap = "The illness-death model.", echo = FALSE}
plot(1, 1, ylim = c(0, 20), xlim = c(0, 30), type = "n", axes = FALSE,
     xlab = "", ylab = "")
rect(1, 12, 9, 18)
text(5, 15, "Healthy")
rect(21, 12, 29, 18)
text(25, 15, "Diseased")
arrows(x0 = 9, y0 = 16, x1 = 21, length = 0.1)
arrows(x0 = 21, y0 = 14, x1 = 9, length = 0.1)
symbols(15.5, 4, circles = 3, add = TRUE, inches = FALSE)
text(15.5, 4, "Dead")
arrows(x0 = 5, y0 = 12, x1 = 12.7, y1 = 5.5, length = 0.1)
arrows(x0 = 25, y0 = 12, x1 = 25 - 6.8, y1 = 5.5, length = 0.1)
```


Individuals may move back and forth between the states *Healthy* and
*Diseased*, and from each of these two states there is a pathway to
*Dead*, which is an *absorbing state*\index{absorbing state},
meaning that once in that state, you never leave
it.\index{illness-death model|)}$\Box$

## More data sets

A few examples and data sets will be used repeatedly throughout the book,
and we give a brief description of them here. They are all available in the
R package **eha**, which is loaded into a running R session by the call
```{r geteha1,echo=TRUE}
library(eha)
```
This loads the **eha** package. In the
examples to follow, we assume that this is already done. The main data
source is the CEDAR, Umeå University, Sweden. However, one
data set is taken from the home page of [Statistics Sweden](http://www.scb.se).

```{example, name = "Survival of males aged 20", label = "menaged"}
```
\index{males aged 20|(}
\index{Data sets!male mortality|(}
This data set is included in the R [@cran] package **eha**
[@eha].
It contains information about 1023 males, age twenty between January 1,
1800 and December 31, 1819, and living in Skellefteå, a parish in the
north-east of Sweden. The total number of records in the data frame
\index{data frame} is 1211, that
is, some individuals are represented by more than one record in the
data file. The reason for that is that the *socio-economic status*
(`ses`) is
one of the covariates in the file, and it changes over time. Each time a
change is recorded, a new record is created for that individual, with the
new value of SES. For instance, the third and fourth rows in the data frame are


```{r mort, echo = FALSE}
print(mort[3:4, ], row.names = FALSE)
```
Note that the variable `id` is the same (3) for the two records, meaning
that both records are information about individual No. 3. The variable
`enter` is age (in years) that has elapsed since the 20th
birth day anniversary, and `exit`
likewise. The information about him is that he
was born on 1800.031, or January 12, 1800, and he is followed from his
21th birth date, or from January 12, 1820. He is in an `upper`
socio-economic status until he is 20 + 13.463 = 33.463 years of age,
when he unfortunately is degraded to a lower `ses`. He is then
followed until 20 years have elapsed, or until his fortieth birthday.  The
variable `event` tells us that he is alive we stop observing him. The
value zero indicates that the follow-up ends with *right censoring*.

In an analysis of male mortality with this data set we could ask whether
there is a socio-economic difference in mortality, and also if it changes
over time. That would typically be done by
*Cox regression*\index{Cox regression} or by a
parametric *proportional hazards model*. More about that follows in
later chapters.
\index{male mortality|)}
\index{Data sets!male mortality|)}

<!---
% \begin{example} {\bf Marital fertility.}\label{ex:1fert}
% This is also a data set available in *eha*, where it is called {\tt fert}.
% It contains lengths of birth intervals for married women in 19th century
% Skellefte{\aa}. The first interval is the time (in years) it took from
% marriage to first birth ({\tt parity} = 0), the second interval (if
% present) is the time elapsed between first and second births ({\tt parity}
% = 1), and so on (compare with the previous discussion of marital
% fertility in this chapter):
% ```{r fert}
% data(fert)
% fert[1:10, ]
% ```
% Here the birth intervals for the two first women in the data file are shown
% (plus the first interval for the third). The first column just shows row
% numbers, then, from left to right, {\tt id} is the key for a mother, {\tt
%   parity} is already explained, {\tt age} is age of mother at the
% *start* of the interval. The first interval of a woman starts at
% marriage, the second at the time of the first birth, and so on. Then {\tt
%   year} is the calendar year at the start of the interval, {next.ivl} is
% the length in years of the (next) interval, {\tt event} is one for all
% interval *except the last}. The variable {\tt prev.ivl} is the length
% of the *previous interval*, if any. This is a variable that could be
% used as an explanatory variable for the length of the next interval. At
% parity 0, there is no previous interval, and that is denoted by {\tt NA},
% meaning {\tt Not Available}, the standard notation for *missing
%   value* in R. In this case, though, the {\tt NA} is not really a missing
% value, but a consequence of definitions.
% Finally, {\tt ses} is short for *socio-economic status* and {\tt
%   parish} is what the name implies. The Skellefte{\aa} region consisted of
% three parishes, *Skellefteå() (by far the largest),
% *Norsjö*, and *Jörn*.

% \end{example}
--->


```{example, name = "Child mortality", label = "childmort1"}
```
\index{mortality!child}
\index{Data sets!child mortality|(}

The data set `child` (in eha) contains follow-up data of children born in 
Skellefteå, northern Sweden, between January 1, 1850 and December 31, 1884. Each 
child is followed a maximum of fifteen years, and the follow-up is interrupted by death,
loss of follow-up (usually out-migration), or the reach of age fifteen.

These data constitute a *cohort* and can be analyzed as such. It may be represented 
in a *Lexis diagram* as in Figure \@ref(fig:lexis21).

```{r lexis21, fig.cap = "Birth cohort 1850-1884, Skellefteå. A child mortality study.", echo = FALSE, fig.height = 4}
plot(1, 1, xlim = c(1850, 1902), ylim = c(0, 17), type = "n", 
     xlab = "Date", ylab = "Age", axes = FALSE)
axis(1, at = c(1850, 1885), labels = c("1850-01-01", "1884-12-31"))
axis(3, at = c(1865, 1900), labels = c("1865-01-01", "1899-12-31"))
axis(2, las = 1)
box()
lines(c(1850, 1850+15), c(0, 15), lty = 2)
lines(c(1885, 1885+15), c(0, 15), lty = 2)
lines(c(1850, 1885), c(0, 0), lty = 2)
lines(c(1850+15, 1885+15), c(15, 15), lty = 2)
text(1875, 7.5, "Cohort data")
```

The first few lines are

```{r cmfirst1, echo = FALSE}
print(child[1:5, colnames(child) != "enter"], row.names = FALSE)
```

Not that the variable `enter` really is redundant here: It is always equal to zero
since all children are followed from birth in Skellefteå.

This data set will be discussed in more detail in a later chapter. It is part of a study of child mortality
\index{Data sets!child mortality|)}$\Box$

```{example, name = "Infant mortality", label = "infmort1"}
```
\index{mortality!infant}
\index{Data sets!infant mortality|(}
This data set is taken from \cite{gb87} and concerns the interplay between
infant and maternal mortality in 19th century Sweden (source: CEDAR,
Umeå University, Sweden). More specifically, we
are interested in
estimating the effect of mother's death on the infant's survival
chances. Because maternal mortality was rare (around one per 200 births),
matching is used. This is performed as follows: for each child experiencing
the death of its mother (before age one), two matched controls were
selected. The criteria were: same age as the case at the event, same sex,
birth year, parish, socio-economic status, marital status of mother. The
triplets so created were followed until
age one, and eventual deaths of the infants were recorded. The data
collected in this way is part of the **eha**  package under the name
  *infants*, and the first rows of the data frame are shown here:

```{r imr1, echo = FALSE}
infa <- infants
##infa$id <- infa$stratum
##infa$stratum <- NULL
names(infa)[1] <- "id"
print(infa[1:3, ], row.names = FALSE)
```
A short description of the variables follows.

*   **id** denotes the id of the mother, 35 in all.
*   **enter** is the age in days of the case, when
its mother died.
*   **exit** is the age in days when
follow-up ends. It takes the value 365 (one year) for those who survived
their first anniversary.
*   **event** indicates whether
a death (1) or a survival (0) was observed.
*   **mother** has value *dead* for all cases and the value *alive* for the
controls.
*   **age** Age of mother at infant's birth.
*   **sex** Sex of the infant.
*   **parish** Birth parish.
*   **civst** Civil status of mother, married or unmarried.
*   **ses** Socio-economic status, often the father's, based on registrations
  of occupation.
*   **year** Calendar year of the birth.


This data set is discussed and analyzed in Chapter 8.
\index{Data sets!infant mortality|)}$\Box$

```{example, name = "Mortality in Sweden 1968-2019, tabular data", label = "swedtab1"}
```
\index{Data sets!tabular}
\index{Data sets!tabular!Swedish mortality|(}
These data sets (`swepop` and `swedeaths`) are taken from Statistics 
Sweden\index{Statistics Sweden}, freely available on
the web site \url{http://www.scb.se}. The aggregated data sets contain yearly
information about population size and number of deaths by sex and age for the
years 1968--2019.

The first and last rows of `swepop` for the year 2019 are

```{r olmscb1, echo = FALSE}
print(swepop[swepop$year == 2019, 1:4][1:3, ], row.names = FALSE)
print(swepop[swepop$year == 2019, 1:4][200:202, ], row.names = FALSE)
```

and the first and last rows of `swedeaths` are

```{r olmscb21, echo = FALSE}
print(swedeaths[swedeaths$year == 2019, 1:4][1:3, ], row.names = FALSE)
print(swedeaths[swedeaths$year == 2019, 1:4][200:202, ], row.names = FALSE)
```



The variables have the following meanings.

*   **pop** Average population size 2007 in the *age* and for the *sex* given on
  the same row. The average is based on the population at the beginning and
  end of the year 2007.
*   **deaths** The observed number of deaths in the *age* and for the
  *sex* given on the same row.
*   **sex** Female or male.
*   **age** Age in completed years.

See Chapter 4 for how to analyze this data set.
\index{Data sets!tabular!Swedish mortality|)}$\Box$
