# Single Sample Data

```{r settings2, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(comment = "", message = FALSE, echo = TRUE, cache = FALSE)
options(width = 60, digits = 7)
```

The basic model descriptions of survival data are introduced. Basically,
the distribution of survival data may
be either _continuous_ or _discrete_, but the nonparametric
estimators\index{nonparametric estimator} of the distributions of survival
time are discrete in any case.
Therefore, the discrete models are
necessary to know in order to understand nonparametric estimation.

In the present chapter, only nonparametric estimation is discussed. This
means that no assumptions whatsoever are made about the true underlying
distribution. Parametric models are presented in Chapter 8.

## Continuous Time Model Descriptions
\index{model!continuous|(}

Traditional statistical model descriptions are based on the
_density_\index{density function}
and the _cumulative distribution_
functions\index{cumulative distribution function}. These functions are not
so suitable for use with censored and/or truncated data. The
_survival_ and _hazard functions_ are better suited, as will be
demonstrated here. It should
however be acknowledged that all these functions are simple functions of
each other, so the information they convey is in
principle the same, it is only convenience that determines which one(s) to
choose and use in a particular application.

### The survival function
\index{survival function|(}

We start with a motivating example (\@ref(exm:sw19)), the life table, and its connection to
death risks and survival probabilities in a population, see 
Table \@ref(tab:lifewomen2).

```{example, sw19, name = "Swedish life tables", echo = TRUE}
```
Statistics Sweden (SCB) is a government agency that produces
statistics and have a coordinating role for the official statistics
of Sweden. From their [home page](https://www.scb.se) it is possible to
download vital
population statistics, like population size and number of deaths by age,
gender, and year. From such data it is possible to construct *life tables*.

Two raw tables are needed, one with population size by age, sex, and year, and one with
the number of deaths by the same categories. Tables are available for age
intervals of lengths 1, 5, and 10 years. We always prefer data with as much
information as possible, so we choose the tables with age interval lengths one. 
It is always possible to *aggregate* later, if that is preferred.

There are two data sets in the **R** package **eha** that contain population size 
information and information about number of deaths, `swepop` and `swedeaths`, 
respectively. Both cover the years 1969--2020. We use both to construct a data frame
`females` containing population size and number of deaths for females 2020. 
Below, the first five and last five rows are printed. Note that the last row, 
with `age = 100` in fact includes all ages from 100 and above. It is common that
tables like this one ends with an open interval.

```{r getpop,echo=TRUE}
library(eha)
females <- swepop[swepop$sex == "women" & swepop$year == 2020, 
                  c("age", "pop")]
females$deaths <- swedeaths[swedeaths$sex == "women" & 
                                swedeaths$year == 2020, "deaths"]
print(females[c(1:5, 97:101), ], row.names = FALSE)
```

A full view of the population size by age is given in Figure \@ref(fig:scbpopfig).

```{r scbpopfig, fig.cap="Female population by age, Sweden 2020.", echo = FALSE}
par(las = 1)
plot(0:100, females$pop / 1000, xlab = "Age", ylab = "Thousands", type = "h",
        axes = FALSE)
axis(1, at = seq(0, 100, by = 20), labels = c(seq(0, 80, by = 20), "100+"))
axis(2)
box()
```

And the numbers of deaths by age are shown in Figure \@ref(fig:scbdeadfig).

```{r scbdeadfig, fig.cap="Female deaths by age, Sweden 2020.", echo = FALSE}
par(las = 1)
plot(0:100, females$deaths / 1000, xlab = "Age", ylab = "Thousands", type = "h", 
        axes = FALSE)
axis(1, at = seq(0, 100, by = 20), labels = c(seq(0, 80, by = 20), "100+"))
axis(2)
box()
```

Our primary interest lies in the *age specific mortality*, which simply is given 
by dividing the number of deaths by the population size, age by age. We do that, 
and add one column, derived from the original three, in the following way.

```{r addtwo, echo = TRUE}
females$risk <- females$deaths / females$pop
alive <- numeric(101)
alive[1] <- 100000
for (i in 2:101){
    alive[i] <- alive[i - 1] * (1 - females$risk[i-1])
}
females$alive <- alive
```
The result is shown in Table \@ref(tab:lifewomen2).

```{r lifewomen2, echo = FALSE}
fem <- females
fem$pop <- round(fem$pop)
fem$risk <- round(fem$risk, 5)
fem$alive <- round(fem$alive)
fem$age <- as.character(format(fem$age, digits = 6))
fem$age[101] <- "100+"
fem$pop <- as.character(fem$pop)
fem$deaths <- as.character(fem$deaths)
fem$risk <- as.character(format(fem$risk, digits = 5))
fem$alive <- as.character(format(fem$alive, digits = 7))
fem[6, ] <- rep("...", 5) # NOTE: Weq
fem <- fem[c(1:5, 97:101), ]
## For text input:
d1 <- round(females$risk[1] * 100000)
l2 <- 100000 - d1
##l2 <- as.character(l2, digits = 5)
##knitr::kable(fem, row.names = FALSE, booktabs = TRUE,
  ##           caption = "Life table for Swedish women 2020.", align = "r")
source("R/tbl.R")
tbl(fem, caption = "Life table for Swedish women 2020.")
```


It is constructed in the following way. The first column, headed by
_age_, contains the ages in completed years. The second column,
headed _pop_ contains the numbers of women in Sweden in the
corresponding ages. It is equal to the mean
population size over the year. Since population numbers from the SCB are
taken at the turn of the years, the numbers in column two are averages of
two columns in the SCB file. The third column contains the total numbers of
deaths in each age category during the year.  The fourth and fifth columns
are derived from the previous ones. The _risk_ is the
ratio between the number of deaths and population size for each age.

Column five, _alive_, contains the actual construction of
the life table. It depicts the actual size of a hypothetical birth cohort\index{cohort},
_year by year_,
exposed to the death risks of column four. It starts with 100000 (an arbitrary, 
but commonly chosen, number) newborn females.
Under the first year of life the death risk is `r fem$risk[1]`, meaning that we
expect `r fem$risk[1]` x 100000 = `r d1` girls to die in the
first year. In other words, 99798 girls will remain alive
at the first anniversary (at exact age one year). That is the number in the
second row in the fifth column. The calculations then continues in the same
way through the last two columns. Note, though, that the last interval
(100+) is an open interval, i.e., it has no fixed upper limit.

Note the difference in interpretation between the numbers in the second
column ("pop") and those in the last ("alive"); the former
gives the age distribution in a population a given year (referred to as
_period data_), while the latter constitutes synthetic 
*cohort data*\index{synthetic cohort}, i.e., a group of newborn are followed from
birth to death, and the
actual size of the (still alive) cohort is recorded age by age. That is,
"individuals" in the synthetic cohort live their whole lives under the
mortality conditions of the year 2020. Of course no real individual can
experience this scenario, hence the name "synthetic
cohort". Nevertheless, it is a useful measure for illustrating the state of
affairs regarding mortality a given year.



```{r lifetablefig, fig.cap = "Life table and survival function, females, Sweden 2020.", fig.scap = "Life table and survival function, females, Sweden 2020.", echo = FALSE,fig.height=4,message=FALSE}
library(eha)
x <- seq(0, 102, length = 1000)
plot(x, ppch(x, 1:100, females$risk, lower.tail = FALSE), type = "l",
     xlab = "Age", ylab = "Surviving fraction", axes = FALSE)
##plot(females$age, females$alive / 100000, type = "s",
  ##   xlab = "Age", ylab = "Surviving fraction",
    ## axes = FALSE)
med <- qpch(0.5, 1:100, females$risk)
axis(1, at = c(0, 50, med, 100), labels = c(0, 50, round(med, 1), 100))
axis(2, at = c(0, 0.5, 1), las = 1)
box()
abline(v = 0)
abline(h = 0)
abline(h = c(0.5, 1), lty = 3)
abline(v = med, lty = 3)
```

It makes sense to plot the life table, see Figure \@ref(fig:lifetablefig). 
Exactly how this is done with individual data will be shown later in the chapter, but here we 
simply plot `alive` (divided by 100000) against `age`. *Median life length* is
`r round(eha::qpch(0.5, 1:100, females$risk), 1)`, see construction in the figure.
By dividing all numbers by the original cohort size (in our case 100000), they
may be interpreted as *probabilities*.\index{life table|)}$\ \Box$

The *survival function* $S(t), \; t > 0$, is defined as
  the probability of surviving past $t$, $t > 0$, or with a formula,

\begin{equation}
S(t) = P(T \ge t), \quad t > 0, 
(\#eq:surv)
\end{equation}

where $T$ is the (random) life length under study. The symbol $P(T \ge t)$
reads *"the probability that $T$ is equal to or larger than $t$"*; 
here $t$ is a fixed number while $T$ denotes a "random
quantity". In statistical language, $T$ is called a *random variable*. 
In our example, $T$ is the (future) life length of a randomly
chosen newborn female, of course unknown at birth.

We will, for the time being, assume that $S$ is a "nice" function, smooth
and *differentiable*. That will ensure that the following definitions
are unambiguous.\index{survival function|)}

### The density function
\index{density function|(}
The *density function* is defined as minus the derivative of the survival function, or

\begin{equation}
f(t) = -\frac{d}{dt} S(t), \quad t > 0
(\#eq:dens)
\end{equation}

The intuitive interpretation of this definition is that, for small enough $s$, 
the following approximation is good:

$$
P(t_0 \le T < t_0 + s) = S(t_0) - S(t_0 + s) \approx s f(t_0). 
$$
This is illustrated in Figure \@ref(fig:densityfig), with $t_0 = 10$ and $s = 2$.

```{r densityfig, fig.cap = "Interpretation of the density function.", echo=FALSE,fig.height=4}
x <- seq(0, 25, length = 100)
f <- function(x) eha::dgompertz(x, scale = 18.9, shape = 1.45, param = "canonical")
y <- f(x)
plot(x[1:85], f(x[1:85]), type = "l", axes = FALSE, ylim = c(-0, max(y) + 0.01), 
     xlab = "Duration.", ylab = "", xlim = c(0, max(x)))
arrows(0, 0, 0, max(y) + 0.01, length = 0.1)
arrows(0, 0, max(x), 0, length = 0.1)
text(max(x), 0.005, "t")
text(0.14, max(y) + 0.01, "f(t)", pos = 4)
ats <- c(0, 10, 12, 20, 30, 40)
axis(1, at = ats, labels = ats)
x0 <- 10
x0s <- 12
axis(2, at = c(0, f(x0), 0.08), labels = round(c(0, f(x0), 0.08), 2), las = 1)
lines(c(x0, x0), c(0, f(x0)), lty = 2)
lines(c(0, x0), (c(f(x0), f(x0))), lty = 2)
lines(c(x0s, x0s), c(0, f(x0s)), lty = 2)
##text(x0, f(x0) + 0.01, expression(paste("P(", t[0], " <= T < ", t[0] + s, ")")), pos = 4)
text(x0, f(x0) + 0.01, "P(10 <= T < 12)", pos = 4)
text(x0, -0.1, expression(t[0]))
text(x0s, -0.1, expression(t[0]+s))
cord.x <- c(x0, seq(x0, x0s, by = 0.01), x0s)
cord.y <- c(0, f(seq(x0, x0s, by = 0.01)), 0)
polygon(cord.x, cord.y, col = "lightgrey")
```

For a short enough interval $(t_0, t_0 +s]$, the probability of an observation falling in that interval is well approximated by the area of a rectangle with sides of lengths $s$ and $f(t_0)$, or $sf(t_0)$. The formal mathematical definition as a limit is given in equation \@ref(eq:denslim).

\begin{equation}
f(t) = \lim_{s \rightarrow 0} \frac{P(t \le T < t + s)}{s}, \quad t > 0.
(\#eq:denslim)
\end{equation}

For the Swedish women 2020, the estimated density function is shown in Figure \@ref(fig:famdens192).

```{r famdens192, echo = FALSE, fig.cap = "Density function for the life length of Swedish women 2020.", ech0 = FALSE}
par(las = 1)
plot(females$age, females$risk * females$alive / 100000, 
     type = "l", xlab = "Age", ylab = "Density")
abline(h = 0)
```


### The hazard function
\index{hazard function|(}

The *hazard function*\index{hazard function} is central to the understanding of survival
analysis, so you are recommended to get at least an intuitive
understanding of it. One way of thinking of it is as an ``instant
probability'' (per unit time); at a given age $t$, it measures the risk of dying in a short
interval $(t, t + s)$ immediately after $t$, for an individual *who
  still is alive at* $t$.

\begin{equation*}
\begin{split}
h(t) &= \lim_{s \rightarrow 0}\frac{P(t \le T < t + s \mid T \ge t)}{s} \\
     &= \lim_{s \rightarrow 0}\frac{S(t) - S(t + s)}{s S(t)} \\
     &= \frac{f(t)}{S(t)}
\end{split}
\end{equation*}


Note the difference between the density and the hazard functions. The
former is (the limit of) an *unconditional* probability, while the
latter is (the limit of) a 
*conditional* probability\index{conditional probability} per time unit. 

For the Swedish women 2020, the estimated hazard function (age-specific mortality) is shown in Figure \@ref(fig:famhaz192).

```{r famhaz192, ech = FALSE, fig.cap = "Hazard function for the life length of Swedish women 2020.", echo = FALSE}
par(las = 1)
plot(females$age, females$risk, type = "l", xlab = "Age", ylab = "Hazard")
abline(h = 0)
```

<!--
Otherwise, the
hazard function is also a kind of density function, with a similar
graphical interpretation, see Figure \@ref(fig:hazardfig) with $t_0 = 10$ and $s = 2$. 

```{r hazardfig, fig.cap = "Interpretation of the hazard function.", echo=FALSE,fig.height=4}
x <- seq(0, 25, length = 100)
f <- function(x) eha::hgompertz(x, scale = 18.9, shape = 1.45, param = "canonical")
y <- f(x)
plot(x[1:85], y[1:85], type = "l", axes = FALSE, xlab ="Remaining years after age 60.",
     ylim = c(0, max(y)), ylab = "", xlim = c(0,max(x)))
arrows(0, 0, 0, max(y), length = 0.1)
arrows(0, 0, max(x), 0, length = 0.1)
text(max(x), 0.02, "t")
text(0.14, max(y), "h(t)", pos = 4)
axis(1, at = ats, labels = ats)
x0 <- 10
x0s <- 12
axis(2, at = round(c(0, f(10), 0.25), 2), las = 1)
lines(c(x0, x0), c(0, f(x0)), lty = 2)
lines(c(x0s, x0s), c(0, f(x0s)), lty = 2)
lines(c(0, x0), c(f(x0), f(x0)), lty = 2)
##text(x0s + 0.1, f(x0s) / 2, expression(paste("P(", t[0], " < T <= ", t[0] + s, " | T >= ", t[0], ")")), pos = 4)
text(x0s + 0.1, f(x0s) / 2, "P(10 <= T < 12  | T >= 10)", pos = 4)
text(x0, -0.1, expression(t[0]))
text(x0s, -0.1, expression(t[0]+s))
cord.x <- c(x0, seq(x0, x0s, by = 0.01), x0s)
cord.y <- c(0, f(seq(x0, x0s, by = 0.01)), 0)
polygon(cord.x, cord.y, col = "lightgrey")
```
-->

### The cumulative hazard function
\index{cumulative hazard function|(} 

The *cumulative hazard function* is
defined as the integral of the hazard function,

\begin{equation*}
H(t) = \int_0^t h(s) ds, \quad t \ge 0.
\end{equation*}
That is, an intuitive interpretation is that the cumulative hazard function
successively accumulates the instant risks.

The cumulative hazard function is important because it is fairly easy to
estimate nonparametrically\index{nonparametric estimation} (i.e., without
any restrictions or assumptions), in contrast to the hazard and density functions. 

For the Swedish women 2020, the estimated cumulative hazard function is shown in Figure \@ref(fig:famcumhaz192).

```{r famcumhaz192, ech = FALSE, fig.cap = "Cumulative hazard function for Swedish women 2020.", echo = FALSE}
par(las = 1)
plot(females$age, cumsum(females$risk), type = "l", xlab = "Age", ylab = "Cumulative hazard")
abline(h = 0)
```

```{example, name = "The exponential distribution"}
```

Perhaps the simplest continuous life length distribution is the
*exponential distribution*. It is simple because its hazard function
is constant:

\begin{equation*}
h(t) = \lambda, \quad \lambda > 0, \; t \ge 0.
\end{equation*}
From this it is easy to calculate the other functions that characterize the
exponential distribution. The cumulative hazards function is

\begin{equation*}
H(t) = \lambda t, \quad \lambda > 0,\; t \ge 0,
\end{equation*}
the survival function is

\begin{equation*}
S(t) = e^{-\lambda t}, \quad \lambda > 0,\; t \ge 0,
\end{equation*}
and the density function is

\begin{equation*}
f(t) = \lambda e^{-\lambda t}, \quad \lambda > 0,\; t \ge 0.
\end{equation*}
The property of constant hazard implies *no aging*. This is not a
realistic property for human mortality, but, as we will see, a useful
benchmark, and a useful tool for modelling mortality over short time
intervals (*piecewise constant hazard*). 
\index{cumulative hazard function|)} 
The exponential distribution is described in detail in 
Appendix \@ref(app:B).\index{model!continuous|)}$\ \Box$

## Discrete Time Models {#disc_2}
\index{model!discrete|(}
So far we have assumed, implicitly or explicitly, that time is
continuous. We will now introduce discrete time survival models, and the
reason is two-fold: (i) Even if data are generated from truly continuous
time models, nonparametric estimation of these models will, as will be
shown later, give rise to estimators corresponding to a discrete time
model. This is an inherent property of nonparametric maximum likelihood
estimation. Thus, in order to study the properties of these estimators, we
need some knowledge of discrete time models. (ii) Data are discrete,
usually through grouping. For instance, life lengths may be measured in
full years, introducing *tied data*\index{tied data}.

It is important to realize that in practice all data are discrete. For
instance, it
is impossible to measure time with infinite precision. Therefore, all data
are more or less rounded. If data are so much rounded that the result is
heavily tied data, true discrete-data models are called for.

Discrete time models will now be introduced. Let $R$ be a discrete random
variable with

*  support\index{support} $(r_1, r_2, \ldots, r_k)$ (positive real numbers, usually $1,
  2, \ldots$ or $0, 1, 2, \ldots$),
*  probability mass function

\begin{equation*}
p_i = P(R = r_i), \quad i = 1, \ldots, k,
\end{equation*}
with $p_i > 0, \quad i = 1, \ldots, k$
and $\sum_{i=1}^k p_i = 1$.

Then
<!--
%\vspace{\baselineskip}
%\begin{center}
%\shadowbox{
%\begin{Beqnarray}
-->

\begin{equation*}
F(t)  =  \sum_{i: r_i \le t} p_i, \quad -\infty < t < \infty,
\end{equation*}

is the cumulative distribution function, and

\begin{equation*}
S(t) = \sum_{i: r_i \ge t} p_i, \quad -\infty < t < \infty,
\end{equation*}

is the survival function.

<!--

% \psset{yunit=1.2cm}

% \begin{figure}
% \pspicture(-1, -1)(10,2)% Density function
% \psaxes[ticks=none,labels=none,linewidth=1pt]{->}(9,1.5)
% \uput[0](-0.7,0){{  0}}
% \uput[0](-0.7,1){{  1}}
% \uput[90](9,-1.5\baselineskip){{ $x$}}
% \uput[90](1,-1.5\baselineskip){ 1}
% \uput[90](4,-1.5\baselineskip){ 4}
% \uput[90](6,-1.5\baselineskip){ 6}
% \uput[90](0.5,1.2){{ $p(x)$}}
% \psline(1,0)(1,0.2)
% \psline(4,0)(4,0.5)
% \psline(6,0)(6,0.3)
% \psline[linewidth=1pt](-0.1,1)(0.1,1) %tick mark on y axis
% %\uput[90](1.5,0.1){\small 0.2}
% %\uput[90](4.5,0.25){\small 0.5}
% %\uput[90](6.5,0.1){\small 0.3}
% \endpspicture
% \caption{Discrete time probability function.}
% \label{fig:discprob}
% \end{figure}

% \begin{figure}
% \pspicture(-1, -1)(10,2)% Survival function
% \psaxes[ticks=none,labels=none,linewidth=1pt]{->}(9,1.5)
% \uput[0](-0.7,0){{  0}}
% \uput[0](-0.7,1){{  1}}
% \uput[90](9,-1.5\baselineskip){{ $x$}}
% \uput[90](1,-1.5\baselineskip){ 1}
% \uput[90](4,-1.5\baselineskip){ 4}
% \uput[90](6,-1.5\baselineskip){ 6}
% \uput[90](0.5,1.2){{ $F(x)$}}
% \psline(0,0)(1,0)
% \psline[linestyle=dotted](1,0)(1,0.2)
% \psline(1,0.2)(4,0.2)
% \psline[linestyle=dotted](4,0.2)(4,0.7)
% \psline(4,0.7)(6,0.7)
% \psline[linestyle=dotted](6,0.7)(6,1)
% \psline(6,1)(9,1)
% \psline[linewidth=1pt](-0.1,1)(0.1,1) %tick mark on y axis
% \endpspicture
% \caption{Discrete time cumulative probability function.}
% \label{fig:disccdf}
% \end{figure}
-->
The discrete time hazard function is defined as

\begin{equation}
h_i = P(R = r_i \mid R \ge r_i) = \frac{p_i}{\sum_{j=i}^k p_j},
\quad i = 1, \ldots, k.
(\#eq:alpha)
\end{equation}

Note that here, the hazard at any given time point is a *conditional
  probability*, so it must always be bounded to lie between zero and
one. In the continuous case, on the other hand, the hazard function may
take any positive value. Further note that if, like here, the support is finite,
the last "hazard atom" is always equal to one (having lived to the "last
station", one is bound to die).

The system \@ref(eq:alpha) of equations has a unique solution, easily
found by recursion:
<!--
%\begin{equation*}
%h_i = \frac{p_i}{1-\sum_{j=1}^{i-1} p_j}, \quad i = 1, \ldots k,
%\end{equation*}
%which has an ``inverse'':
-->

\begin{equation}
p_i = h_i \prod_{j=1}^{i-1} (1 - h_j), \quad i = 1, \ldots, k.
(\#eq:sol)
\end{equation}

From this we get the discrete time survival function at each support point as

\begin{equation*}
S(r_i )  =  \sum_{j = i}^k p_j
         =  \prod_{j = 1}^{i-1} (1-h_j), \quad i = 1, \ldots, k,
\end{equation*}

and the general definition

\begin{equation}
S(t)  =  \prod_{j: r_j < t} \left(1 - h_j\right), \quad t \ge 0
(\#eq:discsur)
\end{equation}

It is easily seen that $S$ is decreasing, $S(0) = 1$, and $S(\infty) = 0$,
as it should be.

```{example, name = "The geometric distribution"}
```
The *geometric* distribution\index{Distributions!geometric|(} has
support on $\{1, 2, \ldots\}$ (another version also includes zero in the
support, this is the case for the one in **R**), and the hazard function $h$ is constant:

\begin{equation*}
h_i = h, \quad 0 < h < 1, \; i = 1, 2, \ldots.
\end{equation*}
Thus, the geometric distribution is the discrete analogue to the
exponential distribution in that it implies no aging. The probability mass
function is, from \eqref{eq:sol},

\begin{equation*}
p_i = h (1-h)^{i-1}, i = 1, 2, \ldots,
\end{equation*}
and the survival function becomes, from \@ref(eq:discsur),

\begin{equation*}
S(t) = (1-h)^{[t]}, \quad t > 0,
\end{equation*}
where $[t]$ denotes the largest integer smaller than or equal to $t$
(rounding towards zero).<!--  In Figure \@ref(fig:geom2) the four functions for a
geometric distribution with $h = 0.25$ are plotted. $\ \Box$
\index{Distributions!geometric|)}


```{r geom2,echo=FALSE,fig.height=5, fig.cap = "The geometric distribution with success probability 0.25.", fig.scap="The geometric distribution.",eval=FALSE}
source("R/plgeo.R")
par(las = 1)
plgeo(h = 0.25, maxval = 10)
```
-->

\index{Distributions!geometric|)}

## Nonparametric Estimators

As an introductory example, look at an extremely simple data set:
$4$, $2^*$, $6$, $1$, $3^*$ (starred observations are right censored; in
the figure, deaths are marked with $+$, censored observations with a small
circle), see Figure \@ref(fig:simple).

```{r simple, fig.cap = "A simple survival data set.",echo=FALSE,fig.height=3}
plot(c(0, 4), c(1, 1), type= "l", ylim = c(0, 5.1), xlim = c(0, 8),
     xlab = "Duration", ylab = "Person No.", axes = FALSE)
axis(1, at = c(0:6))
axis(2, at = 1:5, las = 1)
##box()
abline(v = 0)
text(4.2, 1, "+", cex = 1.2)
lines(c(0, 2), c(2, 2))
text(2, 2, "0")
lines(c(0, 6), c(3, 3))
text(6.2, 3, "+", cex = 1.2)
lines(c(0, 1), c(4, 4))
text(1.2, 4, "+", cex = 1.5)
lines(c(0, 3), c(5, 5))
text(3, 5, "0")
```

<!--
\begin{figure}[ht!]
\psset{yunit=0.6}
\pspicture(-1, -1.5)(10,7)% Data
\psaxes[ticks=none,labels=none,linewidth=1pt]{->}(9,5.5)
\uput[90](9,-1.5\baselineskip){duration}
\uput[90](0.5,5.5){Person No.}
%1
\uput[0](-0.7,1){{\bf  1}}
\psline[linecolor=black](0,1)(4,1)
\uput[0](4.2,1){+}
%\pscircle(4,1){4pt}
%\psline[linestyle=dashed](4,0)(4,5)
%\uput[90](4,-2\baselineskip){$t_{(2)}$}%\psline(0,1)(4,1)
%\uput[90](4,-1.5\baselineskip){ 4}%\psline(0,1)(4,1)
%2
\uput[0](-0.7,2){{\bf  2}}
\psline[linecolor=black](0,2)(2,2)
%\uput[0](4.2,1){+}
\pscircle(2,2){4pt}
%\psline[linestyle=dashed](2,0)(2,5)
%\uput[90](4,-2\baselineskip){$t_{(1)}$}%\psline(0,1)(4,1)
%3
\uput[0](-0.7,3){{\bf  3}}
\psline[linecolor=black](0,3)(6,3)
\uput[0](6.2,3){+}
%\pscircle(4,1){4pt}
%\psline[linestyle=dashed](6,0)(6,5)
%\uput[90](6,-2\baselineskip){$t_{(3)}$}%\psline(0,1)(4,1)
%\uput[90](6,-1.5\baselineskip){ 6}%\psline(0,1)(4,1)
%4
\uput[0](-0.7,4){{\bf  4}}
\psline[linecolor=black](0,4)(1,4)
\uput[0](1.2,4){+}
%\pscircle(4,1){4pt}
%\psline[linestyle=dashed](1,0)(1,5)
%\uput[90](1,-2\baselineskip){$t_{(1)}$}%\psline(0,1)(4,1)
%\uput[90](1,-1.5\baselineskip){ 1}%\psline(0,1)(4,1)
%5
\uput[0](-0.7,5){{\bf  5}}
\psline[linecolor=black](0,5)(3,5)
%\uput[0](4.2,1){+}
\pscircle(3,5){4pt}
%\psline[linestyle=dashed](4,0)(4,5)
%\uput[90](4,-2\baselineskip){$t_{(1)}$}\psline(0,1)(4,1)

\endpspicture
\caption{A simple survival data set.}
\label{fig:simple}
\end{figure}
-->
How should the survival function be estimated?  The answer is that we take
it in 
steps. First, the hazard "atoms" are estimated. It is done nonparametrically,
and the result as such is not very useful. Its potential lies in that it is
used as 
the building block in constructing estimates of the cumulative hazards and
survival functions.

### The hazard atoms

In Figure \@ref(fig:prelhaz), the observed event times in Figure \@ref(fig:simple)
are marked by the vertical dashed lines at durations 1, 4, and 6, respectively. In
the estimation of the hazard atoms, the concept of *risk set* is of vital
importance.

```{r prelhaz, fig.cap = "Preliminaries for estimating the hazard function.",echo=FALSE,fig.height=3}
plot(c(0, 4), c(1, 1), type= "l", ylim = c(0, 5.1), xlim = c(0, 8),
     xlab = "Duration", ylab = "Person No.", axes = FALSE)
axis(1, at = c(0, 1, 4, 6))
axis(2, at = 1:5, las = 1)
abline(v = 0)
text(4.2, 1, "+", cex = 1.2)
lines(c(0, 2), c(2, 2))
text(2, 2, "0")
lines(c(0, 6), c(3, 3))
text(6.2, 3, "+", cex = 1.2)
lines(c(0, 1), c(4, 4))
text(1.2, 4, "+", cex = 1.2)
lines(c(0, 3), c(5, 5))
text(3, 5, "0")
abline(v = c(1, 4, 6), lty = 2)
```

<!--
\begin{figure}
\psset{yunit=0.6cm}
\pspicture(-1, -1.5)(10,6)% Data
\psaxes[ticks=none,labels=none,linewidth=1pt]{->}(9,5.5)
\uput[90](9,-1.5\baselineskip){duration}
\uput[90](0.5,5.5){Person No.}
%1
\uput[0](-0.7,1){{\bf  1}}
\psline[linecolor=black](0,1)(4,1)
\uput[0](4.2,1){+}
%\pscircle(4,1){4pt}
\psline[linestyle=dashed,linecolor=black](4,0)(4,5)
%\uput[90](4,-2\baselineskip){$t_{(2)}$}%\psline(0,1)(4,1)
\uput[90](4,-1.5\baselineskip){ 4}%\psline(0,1)(4,1)
%2
\uput[0](-0.7,2){{\bf  2}}
\psline[linecolor=black](0,2)(2,2)
%\uput[0](4.2,1){+}
\pscircle(2,2){4pt}
%\psline[linestyle=dashed](2,0)(2,5)
%\uput[90](4,-2\baselineskip){$t_{(1)}$}%\psline[linecolor=black](0,1)(4,1)
%3
\uput[0](-0.7,3){{\bf  3}}
\psline[linecolor=black](0,3)(6,3)
\uput[0](6.2,3){+}
%\pscircle(4,1){4pt}
\psline[linestyle=dashed,linecolor=black](6,0)(6,5)
%\uput[90](6,-2\baselineskip){$t_{(3)}$}%\psline(0,1)(4,1)
\uput[90](6,-1.5\baselineskip){ 6}%\psline(0,1)(4,1)
%4
\uput[0](-0.7,4){{\bf  4}}
\psline[linecolor=black](0,4)(1,4)
\uput[0](1.2,4){+}
%\pscircle(4,1){4pt}
\psline[linestyle=dashed,linecolor=black](1,0)(1,5)
%\uput[90](1,-2\baselineskip){$t_{(1)}$}%\psline(0,1)(4,1)
\uput[90](1,-1.5\baselineskip){ 1}%\psline(0,1)(4,1)
%5
\uput[0](-0.7,5){{\bf  5}}
\psline[linecolor=black](0,5)(3,5)
%\uput[0](4.2,1){+}
\pscircle(3,5){4pt}
%\psline[linestyle=dashed](4,0)(4,5)
%\uput[90](4,-2\baselineskip){$t_{(1)}$}\psline(0,1)(4,1)
\endpspicture
\caption{Preliminaries for estimating the hazard function.}
\label{fig:prelhaz}
\end{figure}
-->
The risk set $R(t)$ at duration $t, \; t > 0$ is defined mathematically
as

\begin{equation}
R(t) = \{\text{all individuals under observation at $t-$}\},
(\#eq:rs)
\end{equation}
or in words, the risk set at time $t$ consists of all individuals present and under
observation just prior to $t$. The reason we do not say ``present at time $t$'' is
that it is vital to include those individuals who have an event or are right
censored at the exact time $t$.  In our example, the risk sets $R(1)$, $R(4)$, and
$R(6)$ are the interesting ones. They are:

\begin{equation*}
\begin{split}
R(1) &= \{1, 2, 3, 4, 5\} \\
R(4) &= \{1, 3\} \\
R(6) &= \{3\} \\
\end{split}
\end{equation*}


The estimation of the hazard atoms is simple. First, we assume that the
probability of an event at times where no event is observed, is zero. Then, at
times where events do occur, we count the number of events and divides that
number
by the size of the corresponding risk set. The result is shown in  \@ref(eq:estatoms).

\begin{equation}
\begin{split}
\hat{h}(1) &= \frac{1}{5} = 0.2 \\
\hat{h}(4) &= \frac{1}{2} = 0.5 \\
\hat{h}(6) &= \frac{1}{1} = 1 \\
\end{split}
(\#eq:estatoms)
\end{equation}

See also
Figure \@ref(fig:esthaz).

```{r esthaz, fig.cap = "Nonparametric estimation of the hazard function.", fig.height = 3, echo=FALSE}
par(las = 1)
plot(c(1, 1), c(0, 0.2), type = "l", xlim = c(0, 7), ylim = c(0, 1.2), axes = FALSE, lwd = 1.5, xlab = "Duration", ylab = "")
axis(1, at = c(0, 1, 4, 6))
axis(2, at = c(0, 1))
abline(h = 0, v = 0)
box()
lines(c(4, 4), c(0, 0.5), lwd = 1.5)
lines(c(6, 6), c(0, 1), lwd = 1.5)
text(1.4, 0.2, "0.2")
text(4.4, 0.5, "0.5")
text(6.4, 1.0, "1.0")
```



<!--
\begin{figure}[ht!]
\psset{yunit=2.0cm}
\pspicture(-1, -1)(10,2)% Hazard function
\psaxes[ticks=none,labels=none,linewidth=1pt]{->}(9,1.5)
\uput[0](-0.7,0){{  0}}
\uput[0](-0.7,1){{  1}}
\uput[90](9,-1.5\baselineskip){{ $t$}}
\uput[90](1,-1.5\baselineskip){ 1}
\uput[90](4,-1.5\baselineskip){ 4}
\uput[90](6,-1.5\baselineskip){ 6}
\uput[90](0.5,1.2){{ $\hat{h}(t)$}}
\psset{linecolor=black}
\psline(1,0)(1,0.2)
\psline(4,0)(4,0.5)
\psline(6,0)(6,1)
\psset{linecolor=black}
\psline[linewidth=1pt](-0.1,1)(0.1,1) %tick mark on y axis
\uput[90](1.5,0.1){\small 0.2}
\uput[90](4.5,0.25){\small 0.5}
\uput[90](6.5,0.5){\small 1.0}
\endpspicture
\caption{Nonparametric estimation of the hazard function.}
\label{fig:esthaz}
\end{figure}
-->
As is evident from Figure \@ref(fig:esthaz), the estimated hazard atoms will
be too 
irregular to be of practical use; they need *smoothing*. The simplest
way of 
smoothing them is to calculate the cumulative sums, which leads to the
*Nelson-Aalen* estimator  [@nel72; @aa78] of the cumulative hazards function, see
Section 2.3.2. There are more direct smoothing techniques to get reasonable
estimators of 
the hazard function itself, e.g., kernel estimators
\index{kernel estimator}, but they will not be discussed here. See
e.g. @sil86 for a general introduction to kernel smoothing.

### The Nelson-Aalen estimator
\index{cumulative hazard function|(} 

```{r na, fig.cap="The Nelson-Aalen estimator.", fig.height=4, echo=FALSE}
##library(eha)
##dat <- data.frame(exit = c(4, 2, 6, 1, 3),
##                  event = c(1, 0, 1, 1, 0))
##x <- c(0, 1, 4, 6)
##y <- c(0, 0.2, 0.5, 1.0)
##plot(x, cumsum(y), type = "s", xlim = c(0, 8))
plot(c(0, 1), c(0, 0), type = "l", lwd = 1.5, xlim = c(0, 8), ylim = c(0, 2), xlab = "Duration", ylab = "Nelson-Aalen", axes = FALSE)
axis(1, at = c(0, 1, 4, 6))
axis(2, at = c(0, 0.2, 0.7, 1.7), las = 1)
lines(c(1, 4), c(0.2, 0.2), lwd = 1.5)
lines(c(4, 6), c(0.7, 0.7), lwd = 1.5)
lines(c(6, 8), c(1.7, 1.7), lwd = 1.5)
lines(c(1, 1), c(0, 0.2), lty = 3)
lines(c(4, 4), c(0.2, 0.7), lty = 3)
lines(c(6, 6), c(0.7, 1.7), lty = 3)
abline(h = 0, v = 0)
box()
```

From the theoretical relation we immediately get

\begin{equation*}
\hat{H}(t) =  \sum_{s \le t} \hat{h}(s), \quad t \ge 0,
\end{equation*}

which is the *Nelson-Aalen* estimator  [@nel72; @aa78], see
Figure \@ref(fig:na). The sizes of the jumps are equal to the heights of the
"spikes" in Figure \@ref(fig:esthaz).

<!--
\begin{figure}[ht!]
\psset{yunit=2cm}
\pspicture(-1, -1)(10,2)% Cumulative Hazards function
\psaxes[ticks=none,labels=none,linewidth=1pt]{->}(9,2)
\uput[0](-0.7,0){{  0}}
\uput[0](-0.7,1){{  1}}
\uput[90](9,-1.5\baselineskip){{ $t$}}
\uput[90](1,-1.5\baselineskip){ 1}
\uput[90](4,-1.5\baselineskip){ 4}
\uput[90](6,-1.5\baselineskip){ 6}
\uput[90](0.5,1.2){{ $\hat{H}(t)$}}
\psset{linecolor=black}
\psline(0,0)(1,0)
\psline[linestyle=dotted](1,0)(1,0.2)
\psline(1,0.2)(4,0.2)
\psline[linestyle=dotted](4,0.2)(4,0.7)
\psline(4,0.7)(6,0.7)
\psline[linestyle=dotted](6,0.7)(6,1.7)
\psline(6, 1.7)(9,1.7)
\psset{linecolor=black}
\psline[linewidth=1pt](-0.1,1)(0.1,1) %tick mark on y axis

\endpspicture
\caption{The Nelson-Aalen estimator.}
\label{fig:na}
\end{figure}
-->
\index{cumulative hazard function|)}

### The Kaplan-Meier estimator

From the theoretical relation \@ref(eq:discsur) we get

\begin{equation}
\hat{S}(t) =  \prod_{s < t} \bigl(1 - \hat{h}(s)\bigr), \quad t \ge 0,
(\#eq:km2)
\end{equation}

see also Figure \@ref(fig:km2). Equation \@ref(eq:km2) may be given a
heuristic interpretation: In order to survive time $t$, one must survive
*all* "spikes" (or shocks) that come before time $t$. The
multiplication principle for conditional 
probabilities\index{conditional probability} then gives
equation \@ref(eq:km2).

```{r km2, fig.height = 4, echo = FALSE, fig.cap = "The Kaplan-Meier estimator."}
par(las = 1)
x <- c(0, 1, 4, 6)
y <- c(1, 0.8, 0.8 * 0.5, 0)
plot(c(x[1], x[2]), c(1, 1), xlim = c(0, 8), type = "l", xlab = "Duration", axes = FALSE, lwd = 1.5, ylim = c(0, 1.2), ylab = "Kaplan-Meier")
lines(c(x[2], x[3]), c(y[2], y[2]), lwd = 1.5)
lines(c(x[3], x[4]), c(y[3], y[3]), lwd = 1.5)
lines(c(x[4], 8), c(y[4], y[4]), lwd = 1.5)
axis(1, at = c(0, 1, 4, 6))
axis(2, at = y[4:1])
abline(h = 0, v = 0)
lines(c(1, 1), c(y[1], y[2]), lty = 3)
lines(c(4, 4), c(y[2], y[3]), lty = 3)
lines(c(6, 6), c(y[3], 0), lty = 3)
box()
```

<!--
\begin{figure}[ht!]
\psset{yunit=2cm}
\pspicture(-1, -1)(10,2)% Survival function
\psaxes[ticks=none,labels=none,linewidth=1pt]{->}(9,1.5)
\uput[0](-0.7,0){{  0}}
\uput[0](-0.7,1){{  1}}
\uput[90](9,-1.5\baselineskip){{ $t$}}
\uput[90](1,-1.5\baselineskip){ 1}
\uput[90](4,-1.5\baselineskip){ 4}
\uput[90](6,-1.5\baselineskip){ 6}
\uput[90](0.5,1.2){{ $\hat{S}(t)$}}
\psset{linecolor=black}
\psline(0,1)(1,1)
\psline[linestyle=dotted](1,1)(1,0.8)
\psline(1,0.8)(4,0.8)
\psline[linestyle=dotted](4,0.8)(4,0.4)
\psline(4,0.4)(6,0.4)
\psline[linestyle=dotted](6,0.4)(6,0)
\psline(6,0)(9,0)
\psset{linecolor=black}
\psline[linewidth=1pt](-0.1,1)(0.1,1) %tick mark on y axis

\endpspicture
\caption{The Kaplan-Meier estimator}
\label{fig:km2}
\end{figure}


%\subsubsection{Greenwood's formula}

%\subsection{Life table estimators}
-->

## Doing It in R {#doinR_2}

We show how to do it in **R** by using the male mortality data set `mort`, which is part of the package `eha`. By loading `eha` we have access to `mort`. The **R** function `head` is convenient for looking at a few (six by default) rows in a data frame. 

```{r getmort}
library(eha)
head(mort)
```

This is the normal form of storing left truncated and right censored data. Here, males are followed from their twentieth birthdate until death or their fortieth birthdate, whichever comes first. An indicator for *death* is introduced, called `event`. The value 1
indicates that the corresponding life time is fully observed, while the value
0 indicates a right censored life time. Another common coding scheme is
`TRUE` and `FALSE`, respectively.

### Nonparametric estimation

In the **R** package `eha`,
the `plot` function can be used to plot both Nelson-Aalen (cumulative hazards) and
Kaplan-Meier (survival)
curves. Here is the code: 

```{r esthas,fig=FALSE, eval = FALSE}
par(mfrow = c(1, 2))# Two panels, "one row, two columns".
with(mort, plot(Surv(enter, exit, event), fun = "cumhaz", 
     main = "Cumulativa hazards function",
     xlab = "Duration"))
with(mort, plot(Surv(enter, exit, event),
                main = "Survival function",
                xlab = "Duration"))
```

and the result is seen in Figure \@ref(fig:pshazsur).


```{r pshazsur,fig=TRUE,echo=FALSE,fig.cap="Nelson-Aalen plot (left panel) and Kaplan-Meier plot (right panel), male mortality data.", fig.scap = "Male mortality, Nelson-Aalen and Kaplan-Meier plots.", fig.height = 3}
library(eha)
oldpar <- par(mfrow = c(1, 2), las = 1, cex.main = 0.9) 
##par(las = 1, cex.main = 0.9)
with(mort, plot(Surv(enter, exit, event), fun = "cumhaz",
                main = "Cumulative hazards function",
                xlab = "Duration", conf.int = FALSE))
with(mort, plot(Surv(enter, exit, event), 
                main = "Survival function",
                xlab = "Duration", conf.int = FALSE))
par(oldpar)
```

 Note the use of the
function `with`\index{Functions!\fun{with}}; it tells the `plot`
function that it should get its data `(enter, exit, event)` from the
data frame `mort`. The function `Surv`\index{Functions!\fun{Surv}}, imported
from the package `survival`, creates a "survival object", which is
used in many places. It is for instance *the response* in all
functions that perform regression analysis on survival data.

Note that the "Duration" in Figure \@ref(fig:pshazsur) is duration (in
years) since the day each man became twenty years of age. They are followed until
death or age forty, whichever comes first.  The right hand panel shows
that approximately 25 per cent of the men alive at age twenty died before they
became forty. Data come from a 19th century Swedish sawmill area.

<!--
If the Kaplan-Meier estimator is wanted as a table, use `survfit` in
combination with `coxreg`\index{Functions!coxreg} or 
`coxph`\index{Functions!coxph}. Since the output will have
approximately as many rows as there are distinct event times, we illustrate
it by taking only a small
random sample from `mort`.\index{Functions!sample}
\index{Data sets!male mortality}

```{r rsa2, eval = FALSE}
##library(survival)
indx <- sample(NROW(mort), size = 50, replace = FALSE)
rsa <- mort[indx, ]
fit <- coxph(Surv(enter, exit, event) ~ 1, data = rsa)
s.fit <- survfit(fit)
options(width=70)
summary(s.fit)
```

The functions `coxreg`\index{Functions!coxreg} and
`coxph`\index{Functions!coxph}  both fits *Cox
  regression*\index{Cox regression}
models, which is the topic of the next chapter. Here we fit a regression
model with just a constant term and no covariates. This results in
estimation of the *baseline hazards*\index{baseline hazard function}
which `survfit` transforms to 
elements of the Kaplan-Meier estimator seen above. Note also that, by
default, a 95 per cent confidence band\index{confidence band} is produced around
the estimator. 

You can also plot the output of the fit (`s.fit`) by simply typing

```{r output=FALSE, eval = FALSE}
plot(s.fit)
```

The result is not shown here. It will look similar to the right hand panel
of Figure \@ref(fig:pshazsur). 

-->

### Parametric estimation

It is also possible to fit a parametric model to data with the aid of the
function `phreg`\index{Functions!\fun{phreg}|(} 
(it also works with the
function `aftreg`\index{Functions!\fun{aftreg}}). Just 
fit a "parametric proportional hazards model with no covariates".
<!--
%% Must fix plot.phreg!!
\begin{figure}[ht!]
-->
```{r paramfit2,fig.cap="Male mortality, Weibull fit.", fig.height = 3.2}
par(mfrow = c(1, 2), las = 1)
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = mort)
plot(fit.w, fn = "cum", main = "")
plot(fit.w, fn = "sur", main = "")
```

Note that the default distribution in `phreg` is the *Weibull*
distribution, which means that if no distribution is specified (through the
argument `dist`), then the Weibull distribution is assumed.

<!--
We can also show the parameter estimates. For the interpretations of these,
see Appendix B, the *Weibull* distribution.

```{r showest2, eval = FALSE}
coefficients(fit.w)
```

More about this will be discussed in Chapter 6, Parametric models.
\index{Functions!`phreg`|)} 
-->
