# Poisson Regression

Here *Poisson regression*\index{Poisson regression} is introduced and
its connection to Cox regression\index{Cox regression} is discussed. We
start by defining the Poisson distribution, then discuss the
connection to Cox regression and tabular lifetime data.

## The Poisson Distribution
\index{Distributions!Poisson|(}

The *Poisson distribution* is used for
*count data*, i.e., when the result may be any positive integer *0, 1, 2,
...*, without upper limit. The *probability density function* (pdf)
$P$ of a random variable $X$ 
following a Poisson distribution is

\begin{equation}\label{eq:poisdist}
P(X = k) = \frac{\lambda^k}{k!}\exp(-\lambda), \quad \lambda > 0; \; k = 0,
1, 2, \ldots, 
\end{equation}

The parameter $\lambda$ is both the mean and the variance of the
distribution. In 
Figure \@ref(fig:poisdist) the pdf \eqref{eq:poisdist} is plotted for some
values of $\lambda$.


```{r poisdist,fig=TRUE,echo=FALSE,fig.cap="The Poisson cdf for different values of the mean.", fig.scap="The Poisson cdf."}
oldpar <- par(mfrow = c(2, 2))
lambda <- c(0.5, 1, 5, 25)
x <- list(x0 = 0:5,
          x1 = 0:8,
          x2 = 0:15,
          x3 = 0:45
          )
##for (i in 1:4){
    xx <- x[[1]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[1]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 0.5)))
    xx <- x[[2]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[2]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 1)))
    xx <- x[[3]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[3]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 5)))
    xx <- x[[4]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[4]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 25)))
##}
par(oldpar)
```

Note that when $\lambda$ increases, the distribution looks more and more
like a normal distribution.

In **R**, the Poisson distribution is represented by four functions, `dpois`
  `ppois`, `qpois`, and `rpois`, representing the
  probability density function (pdf), the cumulative distribution function
  (cdf), the quantile function (the inverse of the cdf), and random number
  generation, 
  respectively. See the help page for the Poisson distribution for more
  detail. In fact, this is the scheme present for all probability
  distributions available in **R**.

For example, the upper left bar plot in Figure \@ref(fig:poisdist) is
produced in **R** by
```{r poisfig,fig=FALSE}
barplot(dpois(0:5, lambda = 0.5), axes = FALSE, 
        main = expression(paste(lambda, " = ", 0.5)))
```
\index{Functions!barplot}
Note that the function 
`dpois`\index{Functions!dpois} is *vectorizing*:
```{r poisvec}
dpois(0:5, lambda = 0.5)
```

```{example, name = "Marital fertility"}

As an example where the Poisson distribution may be relevant, we look at
the number of children born to a woman after marriage. The data frame 
*fert* in `eha` can be used to calculate the number of births per married
woman i Skellefteå during the 19th century; however, this data set
contains only marriages with one or more births. Let us instead count the
number of births beyond one.
```{r b2}
library(eha)
f0 <- fert[fert$event == 1, ]
kids <- tapply(f0$id, f0$id, length) - 1
barplot(table(kids))
```

The result is shown in Figure \@ref(fig:b2fig).


```{r b2fig,fig=TRUE,echo=FALSE,fig.cap = "Number of children beyond one for married women with at lest one child.", fig.scap="Number of children beyond one for married women"}
barplot(table(kids)[1:13])
```


The question is: Does this look like a Poisson distribution? One way of
checking this is to plot the *theoretic* distribution with the same
mean (the parameter $\lambda$) as the sample mean in the data.

```{r thpois, eval = FALSE}
lam <- mean(kids)
barplot(dpois(0:12, lambda = lam))
```

The result is shown in Figure \@ref(fig:b2figt).

```{r b2figt,echo=FALSE,fig=TRUE,fig.cap ="Theoretical Poisson distribution."}
lam <- mean(kids)
barplot(dpois(0:12, lambda = lam), names.arg = 0:12)
```


Obviously, the fertility data do not follow the Poisson distribution so
well. It is in fact *over-dispersed* compared to the Poisson
distribution. A simple way to check that is to calculate the sample mean
and variance of the data. If data come from a Poisson distribution, these
numbers should be equal (theoretically) or reasonably close.
```{r meanvar}
mean(kids)
var(kids)
```

They are not very close, which also is obvious from comparing the graphs. $\Box$

## The connection to Cox regression
\index{Cox regression|(}

There is an interesting connection between Cox regression and Poisson
regression, which can be illustrated as follows.

```{r coxpois}
dat <- data.frame(enter = rep(0, 4), exit = 1:4, 
                  event = rep(1, 4), x = c(0, 1, 0, 1))
dat
```

We have generated a very simple data set, `dat`. It consists of four
life lengths with a covariate $x$. A Cox regression, where the covariate
$x$ is related to survival is given by\index{Functions!\coxreg}

```{r coxbb}
library(eha)
fit <- coxreg(Surv(enter, exit, event) ~ x, data = dat)
fit
fit$hazards
```

Note that fit$hazards is a `list` with one component per
stratum. In this case there is only one stratum, so the list has one
component. This component is a matrix with two columns; the first contains
the observed distinct failure times, and the second the corresponding
estimates of the hazard "atoms".

Now, this can be replicated by Poisson regression! First the data set
is summarized around each observed failure time. That is, a snapshot of the
risk set at each failure time is created by the function `toBinary`. 

```{r toB}
library(eha)
datB <- toBinary(dat)
datB
``` 

Note that three new "covariates" are created, `riskset`, `risktime`,
and `orig.row`. In the Poisson regression to follow, the factor 
  `riskset` must be included as a *cluster*. Formally, this is
equivalent to including it as a factor covariate, but for data set with
many distinct failure times, the number of levels will be too large to be
run by `glm`. The function `glmmboot` in `eha` is better suited. 

```{r poisreg}
fit2 <- glmmboot(event ~ x, cluster = riskset, 
                 family = poisson, data = datB)
coefficients(fit2)
``` 

The parameter estimate corresponding to $x$ is exactly the same as in the
Cox regression. The baseline hazard function is estimated with the aid of
the *frailty* estimates for the clustering based on `riskset`.

```{r baserisk}
exp(fit2$frail)
``` 

These numbers are the hazard atoms in the baseline hazard estimation, so
they should be equal to the output from {\tt fit\$hazards} in the Cox
regression. This is not the case, and this depends on the fact that {\tt
  coxreg} estimates the baseline hazards at the mean of the covariate,
i.e., at $x = 0.5$ in our toy example. The easiest way to make the numbers
comparable is
by subtracting 0.5 from $x$ in the Poisson regression.

```{r poisreg2}
datB$x <- datB$x - fit$means
fit2 <- glmmboot(event ~ x, cluster = riskset, 
                 family = poisson, data = datB)
exp(fit2$frail)
```
 
This is exactly what `coxreg` produced.

This correspondence between Poisson and Cox regression was pointed out by
S{\o}ren Johansen @joh83. 
\index{Cox regression|)}

## The connection to the piecewise constant hazards model

For completeness the connection between the Poisson distribution and the
*piecewise constant hazard model*\index{piecewise constant
  hazard}\index{Distributions!piecewise constant hazard}
is mentioned here.  This is explored in detail in Section \@ref(sec:pch6)
(Chapter 6).
