# Poisson Regression

Here *Poisson regression*\index{Poisson regression} is introduced and
its connection to Cox regression\index{Cox regression} is discussed. We
start by defining the Poisson distribution, then discuss the
connection to Cox regression and tabular lifetime data.

## The Poisson Distribution
\index{Distributions!Poisson|(}

The *Poisson distribution* is used for
*count data*, i.e., when the result may be any positive integer *0, 1, 2,
...*, without upper limit. The *probability density function* (pdf)
$P$ of a random variable $X$ 
following a Poisson distribution is

\begin{equation}\label{eq:poisdist}
P(X = k) = \frac{\lambda^k}{k!}\exp(-\lambda), \quad \lambda > 0; \; k = 0,
1, 2, \ldots, 
\end{equation}

The parameter $\lambda$ is both the mean and the variance of the
distribution. In 
Figure \@ref(fig:poisdist) the pdf \eqref{eq:poisdist} is plotted for some
values of $\lambda$.


```{r poisdist,fig=TRUE,echo=FALSE,fig.cap="The Poisson cdf for different values of the mean.", fig.scap="The Poisson cdf."}
oldpar <- par(mfrow = c(2, 2))
lambda <- c(0.5, 1, 5, 25)
x <- list(x0 = 0:5,
          x1 = 0:8,
          x2 = 0:15,
          x3 = 0:45
          )
##for (i in 1:4){
    xx <- x[[1]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[1]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 0.5)))
    xx <- x[[2]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[2]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 1)))
    xx <- x[[3]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[3]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 5)))
    xx <- x[[4]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[4]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 25)))
##}
par(oldpar)
```

Note that when $\lambda$ increases, the distribution looks more and more
like a normal distribution.

In **R**, the Poisson distribution is represented by four functions, `dpois`
  `ppois`, `qpois`, and `rpois`, representing the
  probability density function (pdf), the cumulative distribution function
  (cdf), the quantile function (the inverse of the cdf), and random number
  generation, 
  respectively. See the help page for the Poisson distribution for more
  detail. In fact, this is the scheme present for all probability
  distributions available in **R**.

For example, the upper left bar plot in Figure \@ref(fig:poisdist) is
produced in **R** by
```{r poisfig,fig=FALSE}
barplot(dpois(0:5, lambda = 0.5), axes = FALSE, 
        main = expression(paste(lambda, " = ", 0.5)))
```
\index{Functions!barplot}
Note that the function 
`dpois`\index{Functions!dpois} is *vectorizing*:
```{r poisvec}
dpois(0:5, lambda = 0.5)
```

```{example, name = "Marital fertility"}

As an example where the Poisson distribution may be relevant, we look at
the number of children born to a woman after marriage. The data frame 
*fert* in `eha` can be used to calculate the number of births per married
woman i Skellefteå during the 19th century; however, this data set
contains only marriages with one or more births. Let us instead count the
number of births beyond one.
```{r b2}
library(eha)
f0 <- fert[fert$event == 1, ]
kids <- tapply(f0$id, f0$id, length) - 1
barplot(table(kids))
```

The result is shown in Figure \@ref(fig:b2fig).


```{r b2fig,fig=TRUE,echo=FALSE,fig.cap = "Number of children beyond one for married women with at lest one child.", fig.scap="Number of children beyond one for married women"}
barplot(table(kids)[1:13])
```


The question is: Does this look like a Poisson distribution? One way of
checking this is to plot the *theoretic* distribution with the same
mean (the parameter $\lambda$) as the sample mean in the data.

```{r thpois, eval = FALSE}
lam <- mean(kids)
barplot(dpois(0:12, lambda = lam))
```

The result is shown in Figure \@ref(fig:b2figt).

```{r b2figt,echo=FALSE,fig=TRUE,fig.cap ="Theoretical Poisson distribution."}
lam <- mean(kids)
barplot(dpois(0:12, lambda = lam), names.arg = 0:12)
```


Obviously, the fertility data do not follow the Poisson distribution so
well. It is in fact *over-dispersed* compared to the Poisson
distribution. A simple way to check that is to calculate the sample mean
and variance of the data. If data come from a Poisson distribution, these
numbers should be equal (theoretically) or reasonably close.
```{r meanvar}
mean(kids)
var(kids)
```

They are not very close, which also is obvious from comparing the graphs. $\Box$
