---
output:
  pdf_document: default
  html_document: default
---
# More on Cox Regression

Necessary and vital concepts like
*time-dependent covariates*, *communal covariates*, handling of
*ties*, *model checking*, *sensitivity analysis*, etc, 
are introduced in this chapter.


## Time-varying covariates

Only a special kind of time-varying covariates can be treated in **R** by the
packages `eha` and `survival`, and that is so-called
*piecewise constant* functions. How this is done is best described by
an example.

\begin{example} {Civil status} \end{example}

The covariate (factor) `civil status` (called  $civst$ in the **R** data
frame) is an explanatory variable in a
mortality study, which changes value from 0 to 1 at marriage. How should this be
coded in the data file? The 
solution is to create *two* records (for individuals that marry),
  each with a *fixed* value of $civ\_st$:

1.   Original record: $(t_0, t, d, x(s), t_0 < s \le t)$, married
at time $T$, $t_0 < T < t$:

$$
\text{civst}(s) = \left\{\begin{array}{ll}
             \text{unmarried} , & s < T \\
              \text{married}, & s \ge T
              \end{array}\right.
$$

2.    First new record: $(t_0, T, 0, \text{unmarried})$, *always censored*.

3.    Second new record: $(T, t, d, \text{married})$.


The data file will contain two records like (with $T = 30$) what you can see
in Table \@ref(tab:timevar).

```{r timevar,echo=FALSE}
x <- data.frame(id = c(23, 23),
                enter = c(0, 30),
                exit = c(30, 80),
                event = c(0, 1),
                civst = as.factor(c("unmarried", "married"))
                )
knitr::kable(x, caption = "The coding of a time-varying covariate.")
```


In this way, a time-varying covariate can always be handled by utilizing
left truncation\index{left truncation} and 
right censoring\index{censoring!right}. See also Figure \@ref(fig:civ) for
an illustration.  

```{r civ,fig=TRUE,echo=FALSE,fig.cap="A time-varying covariate (unmarried or married). Right censored and left truncated at age 30 (at marriage).",fig.scap="A time-varying covariate"}
#plot(rnorm(100))
plot(1, 2, type = "n", xlim = c(0, 90), ylim = c(0, 3), xlab = "Age",
     ylab = "", yaxt="n", xaxt = "n")
lines(c(0, 30), c(1,1))
points(30, 1, pch = "c", cex = 1.5)
lines(c(30, 80), c(2, 2))
points(30, 2, pch = "|")
points(83, 2, pch = "+", cex = 1.5)
text(13, 1.2, labels = "civst = unmarried", cex = 1)
text(52, 2.2, labels = "civst = married", cex = 1)
abline(v = 0)
abline(v = 30, lty = 2)
axis(1, at = c(0, 30, 60, 80))
```

And note
that this situation is formally equivalent to a situation with *two
  individuals*, one unmarried and one married. The first one is right
censored at exactly the same time as the second one enters the study (left
truncation).  $\Box$

A word of caution: Marriage status may be interpreted as an
*internal* covariate, i.e., the change of marriage status is
individual, and may depend on health status. For instance, maybe only
healthy persons get married. If so, a vital condition in Cox regression is
violated; the risk of dying may only be modeled by conditions from the
strict history of each individual. Generally, the use of time dependent
covariates is dangerous, and one should always think of possible reversed
causality taking place when allowing for it.

## Communal covariates

Communal\index{communal covariate} (external)\index{external
  covariate} covariates are covariates that vary in time 
outside any individual, and is common to all individuals at each specific
\emph{calendar} time. In econometric literature, such a variable is often
called \emph{exogenous}\index{exogenous covariate}. This could be viewed
upon as a special case of a time-varying covariate, but without the danger
of reversed causality that was discussed above. 

```{example, name = "Did food prices affect mortality in the 19th century?"}

For some 19th century parishes in southern Sweden, yearly time series of
food prices are available. In this example we use deviation from the log
trend in annual rye prices, shown in Figure \@ref(fig:logrye).
```


```{r logrye,fig=TRUE,echo=FALSE,fig.cap = "Log rye price deviations from trend, 19th century southern Sweden.", fig.scap="Log rye price deviations from trend"}
library(eha)
data(logrye)
plot(logrye$year, logrye$foodprices, xlab = "Year", ylab = "Log price",
     type = "l")
abline(h = 0)
```

The idea behind this choice of transformation is to focus on deviations
from the "normal" as a proxy for *short-term variation in economic
  stress*. We illustrate the idea with a subset of the built-in data set `scania`, and we show the information for individual No. 1.

```{r scand,echo=TRUE}
library(eha)
scand <- scania[, c("id", "enter", "exit", "event", "birthdate", "sex", "ses")]
scand[scand$id == 1, ]
```

Now, to put on the food prices as a time-varying covariate, use the function
```make.communal```. We show what happens to individual No. 1.

```{r communal}
scand <- make.communal(scand, logrye[, 2, drop = FALSE], 
                        start = 1801.75)
scand[scand$id == 1, ]
```

A new variable, `foodprices` is added, and each individual's duration
is split up in one year long intervals (except the first and last). This is
the way of treating `foodprices` as a time-varying covariate. 

```{r analcomm}
fit <- coxph(Surv(enter, exit, event) ~ ses + sex + foodprices, 
                data = scand)
round(summary(fit)$coefficients, 3)
```


Socio-economic status and sex do not matter much, but food prices do; the
effect is almost significant at the 5 per cent level. $\Box$


## Tied event times

Tied event times can in theory not occur with continuous-time data, but it
is impossible to measure duration and life lengths with infinite
precision. Data are always more or less rounded, tied event times occur
frequently in practice. This may cause problems (biased estimates) if
occurring too frequently. There are a few ways to handle tied data, and the
so-called exact method considers all possible permutations of the tied
event times in each risk set. It works as shown in the following example.

```{example, name = "Likelihood contribution at tied event time"}

$R_i = \{1, 2, 3\}$, 1 and 2 are events; two
possible orderings:
  
\begin{eqnarray*}
L_i(\boldsymbol{\beta}) & = & \frac{\psi(1)}{\psi(1) + \psi(2) + \psi(3)} \times
            \frac{\psi(2)}{\psi(2) + \psi(3)} \\ & + &
\frac{\psi(2)}{\psi(1) + \psi(2) + \psi(3)} \times
            \frac{\psi(1)}{\psi(1) + \psi(3)} \\
 & = & \frac{\psi(1)\psi(2)}{\psi(1) + \psi(2) + \psi(3)}
 \left\{\frac{1}{\psi(2) + \psi(3)}
            + \frac{1}{\psi(1) + \psi(3)}\right\}
\end{eqnarray*}
$\Box$

The main drawback with the exact method is that it easily becomes
unacceptably slow, because of the huge number of permutations that may be
necessary to consider. It is however available in the function `coxph` in the
`survival` package as an option.

Fortunately, there are a few excellent approximations, most notably
Efron's\index{Efron's approximation}, which is the default method in most
survival packages in 
**R**. Another common approximation, due to Breslow\index{Breslow approximation}, 
is the default in some
statistical software and also possible to choose in the `eha` and 
  `survival` packages in **R**. Finally, there is no cost involved in using
these approximations in the case of no ties at all; they will all give
identical results.

With too much ties in the data, there is always the possibility to use
discrete time methods. One simple way of doing it is to use the option 
`method = "ml"` in
the `coxreg` function in the **R** package `eha`.

```{example, name = "Birth intervals."}
	We look at length of birth intervals
  between first and second births for married women in 19th century
  northern Sweden, a subset of the data set 'fert', available in the
  `eha`   package.  
Four runs with `coxreg` are performed, with all the possible ways of
treating ties. 

```{r 5ties, echo=TRUE}
library(eha)
first <- fert[fert$parity == 1, ]
## Default method (not necessary to specify method = "efron"
fit.e <- coxreg(Surv(next.ivl, event) ~ year + age, data = first, 
                method = "efron")
## Breslow
fit.b <- coxreg(Surv(next.ivl, event) ~ year + age, data = first, 
                method = "breslow")
## The hybrid mppl:
fit.mp <- coxreg(Surv(next.ivl, event) ~ year + age, data = first, 
                 method = "mppl")
## True discrete:
fit.ml <- coxreg(Surv(next.ivl, event) ~ year + age, data = first, 
                 method = "ml")
```

Then we compose a table of the four results, see Table \@ref(tab:5compt). 

```{r 5compt,echo=FALSE}
out <- rbind(fit.e$coefficients, fit.b$coefficients, 
             fit.mp$coefficients, fit.ml$coefficients)
rownames(out) <- c("Efron", "Breslow", "MPPL", "ML")
colnames(out) <- c("year", "age")
knitr::kable(out, booktabs = TRUE, digits = 4, 
      caption = "Comparison of four methods of handling ties, estimated coefficients.")
```


There are almost no difference in results between the four methods. For the
`exact` method mentioned above, the number of permutations is 
$\sim 7 \times 10^{192}$, which is impossible to handle.
$\Box$

With the help of the function `risksets` in `eha`, it is easy to
check the composition of risk sets in general and the frequency of ties in
particular. 
```{r 5rs,echo=TRUE}
rs <- risksets(Surv(first$next.ivl, first$event))
tt <- table(rs$n.event)
tt
```
This output says that `r tt[1]` risk sets have only one event each (no
ties), `r tt[2]` risk sets have exactly two events each, etc.

The object `rs` is a `list` with seven components. Two of them are
`n.event`, which counts the number of events in each risk set, and `size`, 
which gives the size (number of individuals under observation) of
each risk set. Both these vectors are ordered after the `risktimes`,
which is another component of the list. For the rest of the components, see
the help page for `risksets`.

It is easy to produce the *Nelson-Aalen plot*\index{Nelson-Aalen plot}
with the output from `risksets`. The code is
```{r 5nal,fig=FALSE}
plot(rs$risktimes, cumsum(rs$n.event / rs$size), type = "s", 
     xlab = "Duration (years)", ylab = "Cumulative hazards")
abline(h = 0)
```
and the result is shown in Figure \@ref(fig:5nalplot).

```{r 5nalplot,fig=TRUE,echo=FALSE,fig.cap="Nelson-Aalen plot with the aid of the function risksets"}
plot(rs$risktimes, cumsum(rs$n.event / rs$size), type = "s", 
     xlab = "Duration (years)", ylab = "Cumulative hazards")
abline(h = 0)
```

One version of the corresponding survival can be constructed by
```{r 5km,fig=FALSE}
sur <- exp(-cumsum(rs$n.event / rs$size))
plot(rs$risktimes, sur, type = "s", 
     xlab = "Duration (years)", ylab = "Surviving fraction")
abline(h = 0)
```
and the result is shown in Figure \@ref(fig:5kmout)

```{r 5kmout,fig=TRUE,echo=FALSE,fig.cap="Survival plot with the aid of the function risksets.",fig.scap="Survival plot with the aid of the function risksets"}
sur <- exp(-cumsum(rs$n.event / rs$size))
plot(rs$risktimes, sur, type = "s", 
     xlab = "Duration (years)", ylab = "Surviving fraction")
abline(h = 0)
```

	 
