# Poisson Regression

```{r setupcovars5,include=FALSE}
par(las = 1) # Doesn't work? Later?
knitr::opts_chunk$set(comment = "", message = TRUE, echo = TRUE, cache = FALSE)
options(show.signif.stars = FALSE)
library(eha)
```

Here *Poisson regression*\index{Poisson regression} is introduced and
its connection to Cox regression\index{Cox regression} is discussed. We
start by defining the Poisson distribution, then discuss the
connection to Cox regression and tabular lifetime data.

## The Poisson Distribution
\index{Distributions!Poisson|(}

The *Poisson distribution* is used for
*count data*, i.e., when the result may be any positive integer *0, 1, 2,
...*, without upper limit. The *probability density function* (pdf)
$P$ of a random variable $X$ 
following a Poisson distribution is

\begin{equation}\label{eq:poisdist}
P(X = k) = \frac{\lambda^k}{k!}\exp(-\lambda), \quad \lambda > 0; \; k = 0,
1, 2, \ldots, 
\end{equation}

The parameter $\lambda$ is both the mean and the variance of the
distribution. In 
Figure \@ref(fig:poisdist) the pdf \eqref{eq:poisdist} is plotted for some
values of $\lambda$.


```{r poisdist,fig=TRUE,echo=FALSE,fig.cap="The Poisson cdf for different values of the mean.", fig.scap="The Poisson cdf."}
oldpar <- par(mfrow = c(2, 2))
lambda <- c(0.5, 1, 5, 25)
x <- list(x0 = 0:5,
          x1 = 0:8,
          x2 = 0:15,
          x3 = 0:45
          )
##for (i in 1:4){
    xx <- x[[1]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[1]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 0.5)))
    xx <- x[[2]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[2]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 1)))
    xx <- x[[3]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[3]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 5)))
    xx <- x[[4]]
    names(xx) <- xx
##    yy <- lambda[i]
    barplot(dpois(xx, lambda = lambda[4]), axes = FALSE,
##            main = paste(expression(lambda), "=", yy))
            main = expression(paste(lambda, " = ", 25)))
##}
par(oldpar)
```

Note that when $\lambda$ increases, the distribution looks more and more
like a normal distribution.

In **R**, the Poisson distribution is represented by four functions, `dpois`
  `ppois`, `qpois`, and `rpois`, representing the
  probability density function (pdf), the cumulative distribution function
  (cdf), the quantile function (the inverse of the cdf), and random number
  generation, 
  respectively. See the help page for the Poisson distribution for more
  detail. In fact, this is the scheme present for all probability
  distributions available in **R**.

For example, the upper left bar plot in Figure \@ref(fig:poisdist) is
produced in **R** by
```{r poisfig,fig=FALSE,eval=FALSE}
barplot(dpois(0:5, lambda = 0.5), axes = FALSE, 
        main = expression(paste(lambda, " = ", 0.5)))
```
\index{Functions!barplot}
Note that the function 
`dpois`\index{Functions!dpois} is *vectorizing*:
```{r poisvec}
dpois(0:5, lambda = 0.5)
```

```{example, name = "Marital fertility",echo=TRUE}
```
As an example where the Poisson distribution may be relevant, we look at
the number of children born to a woman after marriage. The data frame 
*fert* in `eha` can be used to calculate the number of births per married
woman i Skellefteå during the 19th century; however, this data set
contains only marriages with one or more births. Let us instead count the
number of births beyond one.
The result is shown in Figure \@ref(fig:b2fig).


```{r b2fig,fig=TRUE,echo=FALSE,fig.cap = "Number of children beyond one for married women with at lest one child.", fig.scap="Number of children beyond one for married women"}
f0 <- fert[fert$event == 1, ]
kids <- tapply(f0$id, f0$id, length) - 1
barplot(table(kids)[1:13])
```


The question is: Does this look like a Poisson distribution? One way of
checking this is to plot the *theoretic* distribution with the same
mean (the parameter $\lambda$) as the sample mean in the data.

```{r thpois, eval = FALSE}
lam <- mean(kids)
barplot(dpois(0:12, lambda = lam))
```

The result is shown in Figure \@ref(fig:b2figt).

```{r b2figt,echo=FALSE,fig=TRUE,fig.cap ="Theoretical Poisson distribution."}
lam <- mean(kids)
barplot(dpois(0:12, lambda = lam), names.arg = 0:12)
```


Obviously, the fertility data do not follow the Poisson distribution so
well. It is in fact *over-dispersed* compared to the Poisson
distribution. A simple way to check that is to calculate the sample mean
and variance of the data. If data come from a Poisson distribution, these
numbers should be equal (theoretically) or reasonably close.
```{r meanvar}
mean(kids)
var(kids)
```

They are not very close, which also is obvious from comparing the graphs. $\Box$

## The connection to Cox regression
\index{Cox regression|(}

There is an interesting connection between Cox regression and Poisson
regression, which can be illustrated as follows.

```{r coxpois}
dat <- data.frame(enter = rep(0, 4), exit = 1:4, 
                  event = rep(1, 4), x = c(0, 1, 0, 1))
dat
```

We have generated a very simple data set, `dat`. It consists of four
life lengths with a covariate $x$. A Cox regression, where the covariate
$x$ is related to survival is given by\index{Functions!coxreg}

```{r coxbb}
library(eha)
fit <- coxreg(Surv(enter, exit, event) ~ x, data = dat)
print(summary(fit), short = TRUE)
```
The baseline hazards is given by the function `hazards`: 
```{r gethaz5}
haz <- hazards(fit)
haz
```

Note that `haz` is a *list* with one component per
stratum. In this case there is only one stratum, so the list has one
component. This component is a matrix with two columns; the first contains
the observed distinct failure times, and the second the corresponding
estimates of the hazard "atoms".

Now, this can be replicated by Poisson regression! First the data set
is summarized around each observed failure time. That is, a snapshot of the
risk set at each failure time is created by the function `toBinary`. 

```{r toB}
library(eha)
datB <- toBinary(dat)
datB
``` 

Note that three new "covariates" are created, `riskset`, `risktime`,
and `orig.row`. In the Poisson regression to follow, the variable 
  `riskset` must be included as a *factor*.  The poisson regression is performed
with the `glm` function:

```{r poisreg}
fit2 <- glm(event ~ riskset + x, family = poisson, data = datB)
(co <- coefficients(fit2))
co[2:4] <- co[2:4] + co[1] # Calculate the hazard atoms.
haz.glm <- exp(co[1:4])
``` 

The parameter estimate corresponding to $x$ is exactly the same as in the
Cox regression. The baseline hazard function is estimated with the aid of
the *(Intercept)* and the *riskset* estimates, and we compare them below with 
the hazards estimates from `coxreg`.

```{r baserisk}
xx <- cbind(haz[[1]], haz.glm)
colnames(xx) <- c("Time", "coxreg", "glm")
xx
``` 

Identical results as we can see. This correspondence between Poisson and Cox regression was pointed out by
 @joh83. However, this approach is not very useful in practical work with medium or large
 data sets, because the output from `toBinary` may be huge. For instance, with
 the data set `child`, the resulting data frame consists of slightly more than 
 56 million rows and 11 columns.
 
 **Note:** There is a similar connection to *Binomial   regression*: With the
 `method = "ml"`, we can compare output from `coxreg` and `glm` with `family = binomial`.
 
```{r compacoxml}
fit.ml <- coxreg(Surv(enter, exit, event) ~ x, method = "ml", data = dat)
fit.b <- glm(event ~ riskset + x, family = binomial(link = "cloglog"), data = datB)
rbind(coef(fit.ml),  coef(fit.b)["x"])
```
 
 Identical regression coefficients (and slightly different from what what we got 
 from the Poisson approach). More about discrete time models in Chapter 7, "Parametric models".  
\index{Cox regression|)}

## The connection to the piecewise constant hazard model

For completeness, the connection between the Poisson distribution and the
*piecewise constant hazard model*
\index{piecewise constant hazard}\index{Distributions!piecewise constant hazard}
is mentioned here.  This is explored in detail in Section \@ref(pch6)
(Chapter 7), where the Poisson formulation is cast into a survival analysis
costume and very straightforward to use.

## Tabular lifetime data

Although it is possible to use Poisson regression in place of Cox
regression, the most useful application of Poisson regression in survival
analysis is with tabular data. We have already seen an example of this in 
Chapter 2, the life and its connection to the survival function.

```{example, name = "Mortality in ages 61--80, Sweden 2007."}
```
\label{ex:sw07}
In `eha`, there is the data set `swe07` (a subset and combination of `swepop` 
and `swedeaths`), which contains population size and number of deaths by age 
and sex in Sweden 2007. The age span is restricted to ages 61--80.

```{r swe07}
head(swe07)
tail(swe07)
```

The Poisson model is, where $D_{ij}$ is number of deaths and $P_{ij}$
population size for `age` $i$ and `sex` $j$, $i = 61, \ldots, 80$;
$j = \text{female (0)}, \text{male (1)}$, and $\lambda_{ij}$ is the corresponding
*mortality*, 

\begin{equation*}
D_{ij} \sim \text{Poisson}(\lambda_{ij} P_{ij}), \quad i = 61, 62, \ldots,
80; \; j = 0, 1,
\end{equation*}

with

\begin{equation*}
  \begin{split}
  \lambda_{61,j} P_{61,j} &= P_{61,j}\exp(\gamma + \beta * j) \\
    &= \exp(\log(P_{61,j}) + \gamma + \beta * j), \quad j = 0, 1,
    \end{split}
\end{equation*}

and

\begin{equation*}
  \begin{split}
  \lambda_{ij} P_{ij} &= P_{ij} \exp(\gamma + \alpha_i + \beta * j) \\
  &= \exp(\log(P_{ij}) + \gamma + \alpha_i + \beta * j), \quad i = 62, 63,
  \ldots, 80; \; j = 0, 1.
  \end{split}
\end{equation*}

This calculation shows that it is the log of the population sizes
($\log(P_{ij}$) that is the correct *offset* to use in the Poisson regression.
First we want `age` to be a factor (no restrictions like linearity),
then the **R** function `glm` ("generalized linear model") is used to
fit a Poisson regression model. 

```{r poisreg3}
swe07$age <- factor(swe07$age)
fit <- glm(deaths ~ offset(log.pop) + sex + age, family = poisson, 
           data = swe07)
drop1(fit, test = "Chisq")
```

The function `drop1` is used above to perform two LR tests. The results
are that both variables are highly statistically significant, meaning that
both are very important in describing the variation in mortality over sex
and age. To know *how*, we present (some of) the parameter estimates.

```{r parest}
round(summary(fit)$coefficients[c(1:3), ], 3) # First 3 rows.
round(summary(fit)$coefficients[c(19:21), ], 3) # Last three rows.
```

The results so tell us that males have a distinctly higher mortality than
females, and that mortality steadily increases with age (no
surprises!). The parameter estimates for `age` also opens up the
possibility to simplify the model by assuming a linear or quadratic effect
of age on mortality. We leave that possibility for later. 

One question remains: Is the female advantage relatively the
same over ages? 
To answer that question, we introduce an interaction:

```{r point}
fit1 <- glm(deaths ~ offset(log.pop) + sex * age, 
            family = poisson, data = swe07)
drop1(fit1, test = "Chisq")
```

Note that the `drop1` function only tests the interaction term. This is
because, as we saw before, the main effects tend to be meaningless in the
presence of interaction. However, here there is no sign of interaction, and
we can conclude that in a survival model with `sex` as covariate, we
have *proportional hazards*! This is most easily seen in
*graphs*. The first plot is based on the estimated Poisson model
without interaction. Some calculations are first needed.

```{r modgrcalc}
beta <- coefficients(fit)[2]
alpha <- coefficients(fit)[-2] # Remove sex
alpha[2:length(alpha)] <- alpha[2:length(alpha)] + alpha[1]
lambda.females <- exp(alpha)
lambda.males <- exp(alpha + beta)
```

Then the plot of the hazard functions by

```{r modfig,fig=FALSE}
plot(61:80, lambda.males, ylim = c(0, 0.06), xlab = "Age", 
     type = "s")
lines(61:80, lambda.females, type = "s", lty = 2)
abline(h = 0)
```

Note the parameter `ylim`. It sets the scale on the $y$ axis, and some
trial and error may be necessary to get a good choice.
The function `line`\index{Functions!line} adds a curve to an already
existing plot, and the function `abline`\index{Functions!abline}
adds a straight line; the argument `h = 0` results in a
*horizontal* line. See the help pages for these functions for further
information. 
 
The result is shown in Figure \@ref(fig:poishaz).

```{r poishaz,fig=TRUE,echo=FALSE,fig.cap = "Model based hazard functions for males (upper) and females (lower).", fig.scap = "Hazard functions for males and females, model based."}
plot(61:80, lambda.males, ylim = c(0, 0.06), xlab = "Age", 
     type = "s")
lines(61:80, lambda.females, type = "s", lty = 2)
abline(h = 0)
```

We can also make a plot based only on raw data. For that it is only
necessary to calculate the *occurrence-exposure
  rates*\index{occurrence-exposure rate}. An occurrence-exposure rate is
simply the ratio between the number of occurrences (of some event) divided
by total exposure (waiting) time. In the present example, mean population
size during one year is a very good approximation of the exact total
exposure time (in years).

```{r rawcal}
femal <- swe07[swe07$sex == "female", ]
mal <- swe07[swe07$sex == "male", ]
f.rate <- femal$deaths / femal$pop
m.rate <- mal$deaths / mal$pop
```

And finally the plot of the raw death rates is created as follows,

```{r eval=FALSE, echo=TRUE}
plot(61:80, m.rate, ylim = c(0, 0.06), xlab = "Age", 
     ylab = "Mortality", type = "s")
lines(61:80, f.rate, lty = 2, type = "s")
abline(h = 0)
```

with the result as shown in Figure \@ref(fig:rawplot).

```{r rawplot,fig=TRUE,echo=FALSE, fig.cap = "Hazard functions for males (upper) and females (lower) based on raw data.", fig.scap = "Hazard functions for males and females, raw data"}
plot(61:80, m.rate, ylim = c(0, 0.06), xlab = "Age", 
     ylab = "Mortality", type = "s")
lines(61:80, f.rate, type = "s", lty = 2)
abline(h = 0)
```

The differences between Figure \@ref(fig:poishaz) and
Figure \@ref(fig:rawplot) are very small, showing that the Poisson model
without interaction fits the Swedish old age mortality data very well, at least
in the year 2007.

**Note:** In the life table example in Chapter 2 we came to a different conclusion: It 
was *not* appropriate to assume proportional hazards. The reason for this apparent 
contradiction is that here we are looking at a very short piece of human life span,
from 60 to 80 years of age, and the proportionality assumption fits quite well in 
that short perspective.
$\Box$
\index{Distributions!Poisson|)}
