---
output:
  html_document: default
  pdf_document: default
---
# Multivariate survival models

Sometimes survival data come in clusters, and
multivariate, or *frailty*, models are appropriate to use.

Ever since the paper by @vau79, the concept of
frailty has spread in even wider circles of the research
community. Although their primary purpose was to show various
consequences of admitting individual frailties ("individuals age
faster than cohorts"\index{cohort}, due to the selection effect), the effect was
that people started to implement their frailty model in Cox
regression\index{Cox regression} models.

## An Introductory Example

Let us assume that in a follow-up study, the cohort\index{cohort} is not
homogeneous but instead consists of two equally sized groups with
differing hazard rates. Assume further that we have no indication of
which group an individual belongs to, and that members of both groups
follow an exponential life length distribution:
\begin{equation*}
\begin{split}
h_1(t) &= \lambda_1 \\
h_2(t) &= \lambda_2 \\
\end{split} \qquad t > 0.
\end{equation*}
This implies that the corresponding survival
functions $S_1$ and $S_2$ are
\begin{equation*}
\begin{split}
S_1(t) &= e^{-\lambda_1 t} \\
S_2(t) &= e^{-\lambda_2 t} \\
\end{split} \qquad
t > 0,
\end{equation*}
and a randomly chosen individual will follow the
"population mortality" $S$, which is a *mixture* of the two
distributions:
\begin{equation*}
S(t) = \frac{1}{2} S_1(t) + \frac{1}{2} S_2(t), \quad t > 0.
\end{equation*}
Let us calculate the hazard function for this mixture. We start by
finding the *density function* $f$:
\begin{equation*}
f(t) = -\frac{dS(x)}{dx} = \frac{1}{2}\left(\lambda_1 e^{-\lambda_1 t}
    + \lambda_2 e^{-\lambda_2 t} \right), \quad t > 0.
\end{equation*}
Then, by the definition of $h$ we get
\begin{equation}
h(t) = \frac{f(t)}{S(t)} = \omega(t) \lambda_1 + \big(1 -
\omega(t)\big) \lambda_2, \quad t > 0,
(\#eq:mixhaz)
\end{equation}
with
\begin{equation*}
\omega(t) = \frac{e^{-\lambda_1 t}}{e^{-\lambda_1 t} +
e^{-\lambda_2 t}}
\end{equation*}
It is easy to see that
\begin{equation*}
\omega(t) \rightarrow \left\{ \begin{array}{ll}
                              0, & \lambda_1 > \lambda_2 \\
                              \frac{1}{2}, & \lambda_1 = \lambda_2 \\
                              1, & \lambda_1 < \lambda_2
                             \end{array} \right.
, \quad \mbox{as } t \rightarrow \infty,
\end{equation*}
implying that
\begin{equation*}
h(t) \rightarrow \min(\lambda_1, \lambda_2), \quad t \rightarrow
\infty,
\end{equation*}
see Figure \@ref(fig:fig6frail).

```{r fig6frail,fig=TRUE,echo=FALSE,fig.cap = "Population hazard function (solid line). The dashed lines are the hazard functions of each group, constant at 1 and 2.",fig.scap="Population hazard function."} 
lambda1 <- 1
lambda2 <- 2
w <- function(t) exp(-lambda1 * t) / (exp(-lambda1 * t) + exp(-lambda2 * t))
x <- seq(0, 5, length = 1000)
plot(x, w(x) * lambda1 + (1 - w(x)) * lambda2, ylim = c(0, 2.1), type = "l", xlab = "Time", ylab = "h(t)")
abline(h = 2, lty = 2)
abline(h = 1, lty = 2)
abline(v = 0)
``` 

The important point here is that it is *impossible* to tell from
data alone whether the population is homogeneous, with all
individuals following the same hazard function (\@ref(eq:mixhaz), or
if it in fact consists of two groups, each following a constant
hazard rate.
Therefore, individual frailty models like $h_i(t) = Z_i h(t), \quad i = 1,
\ldots, n$, where $Z_i$ is the "frailty" for individual No.\ $i$,
and $Z_1, \ldots, Z_n$ are independent and identically distributed (iid)
are less useful.

A heuristic explanation to all this is the dynamics of the problem:
We follow a population (cohort) over time, and the *composition*
of it changes over time. The weaker individuals die first, and the
proportion stronger will steadily grow as time goes by.

Another terminology is to distinguish between *individual* and
*population* hazards. In Figure \@ref(fig:fig6frail) the solid line is the
population hazard, and the dashed lines represent the two kinds of
individual hazards present. Of course, in a truly homogeneous population,
these two concepts coincide.

## Frailty models

Frailty models in survival analysis corresponds to *hierarchical*
models in linear or generalized linear models. They are also called
*mixed effects models*. A general theory, with emphasis on using **R**,
of mixed effects models can be found in  @pinbat00.

### The simple frailty model

@vau79 described an individual frailty model,
\begin{equation*}
h(t; \mathbf{x}, Z) = h_0(t) Z e^{\boldsymbol{\beta} \mathbf{x}},
\quad t > 0,
\end{equation*}
where $Z$ is assumed to be drawn independently for each individual.
Hazard rates for ``random survivors''  are not proportional,
but converging (to each other)
if the frailty distribution has finite variance.
Thus, the problem may be less pronounced in AFT than in PH regression.
However, as indicated in the introductory example, with individual frailty
the identification problems are large, and such models are best avoided.

## The shared frailty model

Frailty models work best when there is a natural grouping of the data, so
that observations from the same group are dependent, while two individual
survival times from different groups can be regarded as independent. Such a
model may be described as
\begin{equation}
h_i(t; \mathbf{x}) = h_{i0}(t) e^{\boldsymbol{\beta}\mathbf{x}}, \quad i = 1, \ldots, s;\; t > 0,
(\#eq:strat7)
\end{equation}
which simply is a stratified Cox regression\index{Cox regression} model. By
assuming  
\begin{equation}
h_{i0}(t) = Z_i h_0(t), \quad i = 1, \ldots s;\; t > 0,
(\#eq:frail7)
\end{equation}
the traditional multivariate frailty model emerges. Here it is assumed that
$Z_1, \ldots, Z_s$ are independent and identically distributed
(*iid*), usually with a lognormal distribution. From \@ref(eq:strat7)
and \@ref(eq:frail7) we get, with $U_i = \log(Z_i)$,
\begin{equation*}
h_i(t; \mathbf{x}) = h_{0}(t) e^{\boldsymbol{\beta}\mathbf{x} + U_i}, \quad i = 1, \ldots, s;\; t > 0.
\end{equation*}
In this formulation,  $U_1, \ldots, U_s$ are *iid* normal with mean
zero and unknown variance $\sigma^2$.
Another popular choice of distribution for the $Z$:s is the gamma
distribution.   

In **R**, the package `coxme` [@coxme]\index{Packages!coxme} fits frailty models. 
We look at the fertility data set in the **R** package `eha`.
```{r fert7, message=FALSE} 
library(eha)
head(fert)
``` 
It seems natural to assume that the lengths of birth intervals vary with
mother, so we try a frailty model with `id` as the grouping
variable. Also notice that the first interval for each woman is measured
from marriage (only married women are included in this data set) to first
birth, so we will start by removing them. They are characterized by 
`parity` being 0.
```{r idcoxme7,message=FALSE} 
fe <- fert[fert$parity != 0, ]
library(coxme)
fit <- coxme(Surv(next.ivl, event) ~ age + ses + parity + 
             (1 | id), data = fe)
summary(fit, digits = 3)
``` 
The estimates of the fixed effects have the same interpretation as in
ordinary Cox regression\index{Cox regression}. The question is if the
results point to the 
significance of including frailty terms? In the last line of the output we
get the estimate of the frailty variance, $\sigma^2 = 0.565$, but no $p$-value 
for the test of the null hypothesis $H_0: \sigma = 0$. One explanation to
this is that ordinary asymptotic theory\index{asymptotic theory} does not
hold for parameter values 
at the boundary of the parameter space.

One way to get a feeling for the impact of the frailty effect is to fit the
same model but without frailty, i.e., the term `(1 | id)`.
```{r nofrail7} 
fit0 <- coxreg(Surv(next.ivl, event) ~ age + ses + parity, 
               data = fe)
summary(fit0)
``` 
\index{Functions!\fun{coxreg}}
We can compare the two "max log likelihoods", in the frailty model the
"Integrated" value `r as.integer(round(fit$loglik[2]))`, and in the
fixed effects case 
`r as.integer(round(fit0$loglik[2]))`. The difference is so large 
(`r round(fit$loglik[2] - fit0$loglik[2])`)
that 
we safely can reject the hypothesis that the frailty model is not needed. 
As an "informal" test, you could take twice that difference and treat it as
as a $\chi^2$ statistic with 1 degree of freedom, calculate a $p$-value and
take as real $p$-value one half of that (all this because ordinary
asymptotic theory does not hold for parameter values on the boundary of the
parameter space!). This gives an approximation of the true $p$-value that
is not too bad.

As a final example, let us look at *old age mortality* in the **R** package
`eha`. This example also shows a dangerous situation that is too easy
to overlook. It has nothing to do with frailty, but with a problem caused
by *missing data*\index{missing data}.

```{example name = "Old age mortality for siblings."}
Take a look at the variables in `oldmort`:
```{r oldmortf7} 
head(oldmort)
``` 
The variable `m.id` is *mother's id*. This means that siblings
will have the same value on that variable, and we can check whether we find
a ``sibling effect'' in the sense that siblings tend to have a similar risk
of dying.
```{r oldanal7} 
fit <- coxme(Surv(enter, exit, event) ~ sex + civ + (1 | m.id), 
             data = oldmort)
fit
``` 
Now, compare with the corresponding fixed effects model:
```{r fixold7} 
fit0 <- coxreg(Surv(enter, exit, event) ~ sex + civ, 
               data = oldmort)
summary(fit0)
``` 
Note that we now got a very much smaller value of the maximized log likelihood,
`r as.integer(round(fit0$loglik[2]))` compared to 
`r as.integer(round(fit$loglik[2]))`! Something is wrong, and the big problem is
that the two analyzes 
were performed on different data sets. How is that possible, we used
  `oldmort` on both occasions? The variable `m.id` has a lot of missing
values, almost 50 per cent are missing (NA), and the standard treatment of NA:s in
**R** is to simply remove each record that contains an NA on any of the
variables in the analysis. So, in the first case, the frailty model, a lot
of records are removed before analysis, but not in the second. To be able
to compare the models we must remove all records with `m.id = NA` from
the second analysis.
```{r shrink7}
olm <- oldmort[!is.na(oldmort$m.id), ]
fit0 <- coxreg(Surv(enter, exit, event) ~ sex + civ, 
               data = olm)
summary(fit0)
```
This is another story! We now got  very similar values of the maximized log likelihoods,
`r round(fit0$loglik[2], 2)` compared to 
`r round(fit$loglik[2], 2)`! The conclusion is that in this case, there
is no frailty effect whatsoever.

One lesson to learn from this example is that you have to be very cautious
when a data set contains missing values. Some functions, like `drop1`,
give a warning when a situation like this is detected, but especially when
comparisons are made in more than on step, it is too easy to forget the
dangers. 

Also note the warning that is printed in the results of 
  `coxme`\index{Functions!\fun{coxme}}: 
  *3163 observations deleted due to missingness*. This is a warning that
should be taken seriously.
$\Box$

<!--
%\end{example}
% \subsubsection{Swedish marriages 1970}
%   From SCB, distribution of {\blue time to divorce}.

% ```{r echo=FALSE,fig=TRUE} 
% source("div.R")
% p <- div(fig = FALSE)
% F <- cumsum(c(0, p))
% plot(0:20, 100 * F, type = "l", xlab = "Years after marriage", ylab = "Divorced fraction (per cent)", col = "blue")
% abline(h = 0)
% abline(v = 0)
% ```

% ```{r echo=FALSE,fig=TRUE} 
% source("div.R")
% p <- div(fig = FALSE)
% F <- cumsum(c(0, p))
% plot(0:20, 100 * (1 - F), type = "l", xlab = "Years after marriage", ylab = "Still married fraction (per cent)", ylim = c(0, 100), col = "blue")
% abline(h = 0)
% abline(v = 0)
% ```

% {\blue Note:} Assumes no deaths.
% ```{r echo=FALSE,fig=TRUE} 
% ##source("div.R")
% p <- div(fig = FALSE)
% F <- cumsum(c(0, p))
% plot(0:20, 100 * (1 - F), type = "l", xlab = "Years after marriage", ylab = "Still married fraction (per cent)", ylim = c(0, 100), col = "blue")
% abline(h = 0)
% abline(v = 0)
% ```

% {\blue Note:} Assumes no deaths.
% \subsubsection{Marriages, a simulation (hazards)}

% ```{r fig=TRUE,echo=FALSE} 
% source("simdiv.R")
% simdiv(what = "haz")
% ```

% ```{r fig=TRUE,echo=FALSE} 
% ##source("R/simdiv.R")
% simdiv(what = "cdf")
% ```
-->

## Parametric frailty models

<!--
% In \R\ there is a package called {\tt lme4}\index{lme4} \citep{lme411},
% which is a de 
% facto standard for analyzing mixed effects models, at least in the
% \R\ community. In that package there is a function {\tt glmer} that
% analyzes generalized linear mixed models (GLMM), and it is possible to
% It is possible utilize the connection between Poisson\index{Distributions!Poisson}
% regression and  the piecewise constant proportional hazards model
% \index{Distributions!piecewise constant hazard}
% discussed in Chapter 6 to fit parametric frailty models. We look at the
% fertility data again.  
% ```{r fert7par} 
% fe13 <- survSplit(fe, end = "next.ivl", event = "event", 
%                   cut = 1:13, episode = "years", start = "start")
% fe13$years <- as.factor(fe13$years) # 1
% fe13$offs <- log(fe13$next.ivl - fe13$start) # 2
% fit <- glmmML(event ~ age + year + ses + years + offset(offs), cluster = id, 
%             data = fe13, family = poisson)
% drop1(fit, test = "Chisq")
% summary(fit0)
% ##logLik(fit0)
% ``` 
% This is the model without taking into account that there is a possible
% dependency structure in the data due to the repeated measurements; most
% mothers have several births and birth intervals. Now the mixed effects approach.
% ```{r poisfert7} 
% library(lme4)
% fit1 <- glmer(event ~ age + year + ses + years + offset(offs) + (1 | id), 
%               family = poisson, data = fe13, nAGQ = 8)
% print(fit1)
% ``` 
% The function \fun{glmer} gives a lot of output, most notably the
% correlation matrix for the fixed effects estimates. All the years makes the
% output just too much. In this example the clustering effect is very small.  
% If we want to test the presence of a random effect over women, an
% alternative is to use \fun{glmmML} in \pkg{eha}.
% ```{r poisfertML7} 
% fit2 <- glmmML(event ~ age + year + ses + years + offset(offs), cluster = id, 
%               family = poisson, method = "ghq", data = fe13, n.points = 8)
% print(fit2)
% ``` 
-->

It is possible utilize the connection between Poisson\index{Distributions!Poisson}
regression and  the piecewise constant proportional hazards model
\index{Distributions!piecewise constant hazard}
discussed in Chapter 6 to fit parametric frailty models. We look at the
fertility data again.  The standard analysis without frailty effects is
```{r stanal7} 
fit0 <- phreg(Surv(next.ivl, event) ~ parity + ses,
              dist = "pch", cuts = 1:13, data = fe)
summary(fit0)
``` 

For testing the presence of a random effect over women, an
alternative is to use `glmmML`\index{Functions!\fun{glmmML}} in `eha`.
```{r poisfertML7} 
library(glmmML)
fe13 <- survSplit(fe, end = "next.ivl", event = "event", 
                  cut = 1:13, episode = "years", 
                  start = "start") #1
fe13$years <- as.factor(fe13$years) # 2
fe13$offs <- log(fe13$next.ivl - fe13$start) # 3
fit1 <- glmmML(event ~ parity + ses + years + offset(offs), 
               cluster = id, family = poisson, method = "ghq", 
               data = fe13, n.points = 9)
summary(fit)
``` 
Note the use of the function `survSplit` (from the `survival` package) (\# 1), the transformation to
`factor` for the slices of time (`years`) created by
`survSplit`, and the creation of an offset (\# 3). This is described in
detail in Chapter 6.

The conclusion is very much the same as with the nonparametric
(`coxme`) approach: The clustering effect of *mother* is very
strong and must be taken into account in the analysis of birth intervals.
The nonparametric approach is easier to use and recommended, but see the next session. 



## Stratification
\index{stratification}
A simple way to eliminate the effect of clustering is to *stratify* on
the clusters. In the birth intervals example, it would mean that intervals
are only compared to other birth intervals from the same mother. The
drawback with a stratified analysis is that it is not possible to estimate
the effect of covariates that are constant within clusters. In the birth
intervals case, it is probable that `ses`, socio-economic status, would
vary little within families. On the other hand, the effet of `birth
  order` or `mother's age` would be suitable to analyze in a stratified
setting. 
```{r strat7} 
library(glmmML)
fit2 <- coxreg(Surv(next.ivl, event) ~ parity + prev.ivl + 
               strata(id), data = fe)
summary(fit2)
``` 
Contrast this result with an unstratified analysis.
```{r unstrat7} 
fit3 <- coxreg(Surv(next.ivl, event) ~ parity + prev.ivl, 
               data = fe)
summary(fit3)
``` 
Note how the effect of `parity` is diminished when aggregating
comparison over all women, while the effect of the length of the previous
interval is enlarged. Try to explain why this result is expected! 

```{example, name =  "Matched data."}

Under certain circumstances it is actually possible to estimate an
effect of a covariate that is constant within strata, but only if it is
interacted with a covariate that is not constant within strata. 
See Example \@ref(exm:matchex9) in Chapter 9.
