# (APPENDIX) Appendix {-}

# Basic statistical concepts

The statistical concepts that are important for understanding what is
going on in this book are gathered here, but briefly treated. The
interested reader who wants a deeper understanding of statistical concepts
should have no problems in finding suitable text books. There are for
instance some recent texts that teaches statistics and how to use it in
**R** [@pd08]. 

## Statistical inference

Statistical inference is the science that help us draw conclusions about
real world phenomena by observing and analyzing samples from them. The
theory rests on probability theory and the concept of *random*
sampling. The statistical analysis never gives absolute truths, but only
statements coupled to certain measures of their validity. These measures
are almost always *probability statements*.

The crucial concept is that of a *model*, despite the fact that the
present trend in statistical inference is towards nonparametric
statistics. It is often stated that with today's huge data taster's,
statistical models are unnecessary, but nothing could be more wrong. 

The important idea in a statistical model is the concept of a
*parameter*. It is often confused with its estimator from data. For
instance, when we talk about *mortality* in a population, it is a
hypothetical concept that is different from the ratio between the observed
number of deaths and the population size (or any other measure based on
data). The latter is an *estimate* (at best) of the former. The whole
idea about statistical inference is to extract information about a
*population parameter* from observing data.

### Point estimation
The case in *point estimation* is to find the best guess (in some
sense) of a population parameter from data. That is, we try to find the
best single value that is closest to the true, but unknown, value of the
population parameter.

Of course, a point estimator is useless if it is not connected to some
measure of its uncertainty. That takes us to the concept of *interval
  estimation. 
  
### Interval estimation
\index{confidence interval|(}\index{interval estimation|(}
The philosophy behind *interval estimation* is that a guess on  a
single value of the unknown population parameter is useless without an
accompanying measure of the uncertainty of that guess. A 
*confidence interval* is an interval, in which we say that the true value of the
population parameter lies with a certain probability (often 95 per cent).
\index{confidence interval|)}\index{interval estimation|)}

### Hypothesis testing

We are often interested in a specific value of a parameter, and in
regression problems this value is almost always *zero* (0). The reason
is that regression parameters measure *effects*, and to test for an
effect is then equivalent to testing that the corresponding parameter has
value zero.

There is a link between interval estimation and hypothesis testing: To test
the hypothesis that a parameter value is zero can be done through
constructing a confidence interval for the parameter. The test rule is
then: If the interval does not cover *zero*, reject the hypothesis,
otherwise do not.

#### The log-rank test {#logrank_A}

The general hypothesis testing theory behind the log-rank test builds on
the *hyper-geometric distribution*. The calculations under the null
hypothesis of no difference in survival chances between the two groups are
performed 
*conditional on both margins*\index{conditional probability}. In
Table \@ref(tab:lrgen), if  
the margins are fixed there is only one degree of freedom left; for a given
value of (say) $d_1$, the three values $d_2$, $(n_1 - d_1)$, and $(n_2 - d_2)$ are determined.

```{r lrgen, echo=FALSE}
tt <- data.frame(Group = c("I", "II", "Total"), 
                 Deaths = c("d1", "d2", "d"), 
                 Survivors = c("n1 - d1", "n2 - d2", "n - d"), 
                 Total = c("n1", "n2", "n"))
knitr::kable(tt, caption = "General table in a log rank test.", booktabs = TRUE,
             row.names = FALSE)
```


Utilizing the fact that, under the null, $d_1$ is hyper-geometrically
distributed, results in the following algorithm for calculating a test
statistic as follows:


1.  Observe $O = d_1$
2.  Calculate the expected value $E$ of $O$ (under the null): 

  $$
  E = d \frac{n_1}{n}.
  $$
3.  Calculate the variance $V$ of $O$ (under the null): 

$$V = \frac{(n - d) d n_1 n_2}{n^2 (n - 1)}.$$
4.  Repeat 1. - 3. for all tables and aggregate according to
  equation \@ref(eq:lorat).

The log rank test statistic $T$ is

\begin{equation}
T = \frac{\sum_{i=1}^k \left(O_i -
E_i\right)}{\sqrt{\sum_{i=1}^k V_i}}
 (\#eq:lorat)
\end{equation}

Note carefully that this procedure is *not* equivalent to aggregating all
tables of raw data!

Properties of the log rank test;

1.  The test statistic $T^2$ is approximately distributed as $\chi^2(1)$.
2.  It is available in most statistical software.
3.  It can be generalized to comparisons of more than two groups.
4.  For $s$ groups, the test statistic is approximately $\chi^2(s-1)$.
5.  The test has *high power* against alternatives with
  *proportional hazards*, but can be weak against non-pro\-por\-tional
  alternatives.

## Presentation of results

The presentation of the results of fitting a regression model to given data is usually
done both graphically and in tabular form. Some general guidelines that must be 
followed are given here. In order to make the presentation easier to understand,
a real data, but fictive, research project is followed.

### The project

In a mortality study of the small town of Umeå in twentieth century northern Sweden we
are especially interested in investigating the mortality among persons aged 50 and above.
We are especially interested in whether there were any differences mortality
between social branches. For this purpose, the data set `ume` (available in
the package **eha**), with selected variables, is used. A summary of the data:

```{r creA, echo = FALSE, message = FALSE}
library(skum)
ume <- obs[obs$region == "ume", ]
ume$event <- ume$sluttyp == 2
ume <- ume[, c("id", "sex", "birthdate", "enter", "exit", "event", "civst", "socBranch")]
ume <- age.window(ume, c(50, 100))
ume <- ume[!is.na(ume$socBranch), ]
levels(ume$civst) <- c("unmarried", "married", "prev.married", "prev.married", "prev.married")
ume$civst <- relevel(ume$civst, ref = "married")
ume$cohort <- cut(ume$birthdate, c(1810, 1858, 1875, 1890, 1921), dig.lab = 4, right = FALSE)
ume$exit[ume$enter >= ume$exit] <- ume$enter[ume$enter >= ume$exit] + 0.001
##ume <- ume[ume$cohort == "[1810,1858)", ]
##ume$cohort <- NULL
umetab <- toTpch(Surv(enter, exit, event) ~ sex + cohort + civst + socBranch,
                 data = ume, cuts = seq(50, 95, by = 3))
##summary(ume)
head(umetab)
```

As you can see, data is *tabulated*, so a piecewise constant hazard model is 
called for. 

### Tabular presentation

We fit a simple model with only one ccovariate, but select only 
*married men*.  This can be done with the help of *Poisson regression* and the 
`glm` function:

```{r glmA}
fit.glm <- glm(event ~ offset(log(exposure)) + age + socBranch, 
               data = umetab, subset = sex == "male" & civst == "married", 
               family = "poisson")
summary(fit.glm)
```
Besides the slightly disturbing fact that all age estimates are printed, the 
presentation of the result for the covariate of interest, `socBranch`, is less
satisfactory from a "survival analysis" point of view: We are generally not 
interested of looking at the estimated baseline in tabular form (we want a
graph). There is a more general critique of the presentation: Too many *p*-values 
are given, and not the relevant ones either.

In `eha`, the function `tpchreg` does the same thing, but with a more satisfactory
presentation of the result:


```{r tabcA}
fit <- tpchreg(oe(event, exposure) ~ socBranch, 
               data = umetab[umetab$civst == "married" & umetab$sex == "male", ], 
               time = age)
summary(fit)
```

First, we got rid of the numerous baseline hazards estimates, and second, the 
*factor* `socBranch` is presented in full, with all four levels. And finally, 
the *relevant p-value*. Because, first,  the truth is that *Wald* p-values are 
notoriously unreliably in nonlinear regression, and second, we want *p*-values for
*covariates* and *not* for *levels* ov covariates.

However, note that the essential findings with the two approaches are identical.

### Graphics

The baseline hazard function is shown in Figure \@ref(fig:bhfA).

```{r bhfA, fig.cap = "The cum"}
plot(fit)
```




## Asymptotic theory
\index{asymptotic theory|(}

### Partial likelihood

Here is a very brief summary of the asymptotics concerning the partial
likelihood. Once defined, it turns out that you may treat it as an ordinary
likelihood function \cite{abgk93}. The setup is as follows.

Let $t_{(1)}, t_{(2)}, \ldots, t_{(k)}$ the ordered observed event times
and let $R_i = R(t_{(i)})$ be the risk set at $t_{(i)}, \; i = 1,
\ldots, k$, see equation \eqref{eq:rs}.
At $t_{(i)}$, *condition* with respect to the composition of $R_i$ and
that one event occurred (for tied events, a correction is necessary).

Then the contribution to the partial likelihood from $t_{(i)}$ is

\begin{multline*}
L_i(\boldsymbol{\beta})  =  P(\mbox{No. $m_i$ dies} \mid
\mbox{one event occur}, R_i) \\
=  \frac{h_0(t_{(i)}) \exp(\boldsymbol{\beta} \mathbf{x}_{m_i})}
{\sum_{\ell \in R_i}h_0(t_{(i)})\exp(\boldsymbol{\beta} \mathbf{x}_\ell)}
=  \frac{\exp(\boldsymbol{\beta} \mathbf{x}_{m_i})}
{\sum_{\ell \in R_i}\exp(\boldsymbol{\beta} \mathbf{x}_\ell)}
\end{multline*}
and the full partial likelihood is

\begin{equation*}
L(\boldsymbol{\beta}) = \prod_{i=1}^k L_i(\boldsymbol{\beta})
 = \prod_{i=1}^k \frac{ \exp(\boldsymbol{\beta} \mathbf{x}_{m_i})}
{\sum_{\ell \in R_i}\exp(\boldsymbol{\beta} \mathbf{x}_\ell)}
\end{equation*}
This is where the doubt about the partial likelihood comes in; the
conditional probabilities\index{conditional probability} multiplied
together do not have a proper 
interpretation as a conditional probability. Nevertheless, it is prudent to
proceed as if the expression really is a likelihood function. The log
partial likelihood becomes

\begin{equation}\label{eq:logplA}
\log\big(L(\boldsymbol{\beta})\big) = \sum_{i=1}^k \left\{\boldsymbol{\beta} \mathbf{x}_{m_i} -
\log\left(\sum_{\ell \in R_i} \exp(\boldsymbol{\beta} \mathbf{x}_\ell)\right)\right\},
\end{equation}
and the components of the *score vector*\index{score vector} are

\begin{equation}\label{eq:scoreA}
\frac{\partial}{\partial \beta_j} \log L(\boldsymbol{\beta})  =
\sum_{i=1}^k \mathbf{x}_{m_i j} - \sum_{i=1}^k \frac{\sum_{\ell \in R_i}
x_{\ell j} \exp(\boldsymbol{\beta} \mathbf{x}_\ell)} {\sum_{\ell \in R_i}
\exp(\boldsymbol{\beta} \mathbf{x}_\ell)}, \quad j = 1, \ldots, s.
\end{equation}
The *maximum partial likelihood (MPL)* estimator of
$\boldsymbol{\beta}$, $\hat{\boldsymbol{\beta}}$, is found by setting \eqref{eq:scoreA} equal to
zero and solve for $\boldsymbol{\beta}$.

For inference, we need to calculate
the inverse of minus the *Hessian*, evaluated at $\hat{\boldsymbol{\beta}}$. This
gives the
estimated *covariance matrix*.
The Hessian is the matrix of the second partial derivatives. The
expectation of minus the Hessian is called the *information matrix*.
The *observed* information matrix is

\begin{equation*}
\hat{I}(\hat{\boldsymbol{\beta}})_{j,m} = -\frac{\partial^2 \log L(\boldsymbol{\beta})}
{\partial \beta_j \partial \beta_m}\mid_{\boldsymbol{\beta} = \hat{\boldsymbol{\beta}}}
\end{equation*}
and asymptotic theory says that

\begin{equation*}
  \hat{\boldsymbol{\beta}} \sim N(\boldsymbol{\beta}, \hat{I}^{-1}(\hat{\boldsymbol{\beta}}))
\end{equation*}
This is to say that $\hat{\boldsymbol{\beta}}$ is asymptotically unbiased and normally
distributed with the given covariance matrix (or the limit of it).
Further, $\hat{\boldsymbol{\beta}}$ is a consistent estimator of $\boldsymbol{\beta}$.
These results are used for for hypothesis testing, confidence intervals, and variable selection.

Note that these are only *asymptotic* results, i.e., useful in
*large to medium sized samples*.
In small samples, *bootstrapping* is a possibility. This option is
available in the **R** package **eha**.

Here a warning is in order: Tests based on standard errors (*Wald*)
tests\index{Tests!Wald})
may be *highly unreliable*, as in all *non-linear* regression
\cite{hd77}. A better alternative is the *likelihood ratio
test*\index{Tests!likelihood ratio}.
\index{asymptotic theory|)}

## Model selection {#modsel_A}

In regression models, there is often several competing models for
describing data. In general, there are no strict rules for ``correct
selection''. However, for *nested* models\index{model!nested}, there
are some formal guidelines. For a precise definition of this concept, see
Appendix A. 

### Comparing nested models

The meaning of *nesting* of models is best described by an example. 

```{example name = "Two competing models"}
```

1.   ${\cal M}_2:\; h(t; (x_1, x_2)) = h_0(t) \exp(\beta_1 x_1 +
     \beta_2 x_2)$
2.    ${\cal M}_1:\; h(t; (x_1, x_2)) = h_0(t) \exp(\beta_1 x_1)$:
      $x_2$ has no effect.

Thus, the model ${\cal M}_1$ is a special case of ${\cal M}_2$ ($\beta_2 = 0$).
We say that ${\cal M}_1$ is *nested* in ${\cal M}_2$.
Now,  *assume* that ${\cal M}_2$ is *true*. Then, 
testing the hypothesis $H_0: \; {\cal M}_1$ is true (as well)
is the same as testing the hypothesis $H_0;\; \beta_2 = 0$. 

The formal theory for and procedure for performing the likelihood ratio
test (LRT)\index{Tests!likelihood ratio} can be summarized as follows:

1.  Maximize $\log L(\beta_1, \beta_2)$ under ${\cal M}_2$; gives $\log
    L(\hat{\beta}_1, \hat{\beta}_2)$.

2.  Maximize $\log L(\beta_1, \beta_2)$ under ${\cal M}_1$, that
    is, maximize $\log L(\beta_1, 0)$; gives $\log L(\beta_1^*, 0)$.

3.  Calculate the test statistic
     \begin{equation*}
     T = 2\big(\log L(\hat{\beta}_1, \hat{\beta}_2) - \log L(\beta_1^*,
     0)\big)
     \end{equation*}

4.   Under $H_0$, $T$ has a $\chi^2$ (chi-square)
     distribution with $d$ degrees of freedom: $T \sim \chi^2(d)$,
     where $d$ is the difference in numbers of parameters in the two
     competing models, in this case $2-1=1$.

5.   Reject $H_0$ if $T$ is large enough. Exactly how much that is depends
     on the level of significance; if it is $\alpha$, choose the limit $t_d$
     equal to the $100 (1 - \alpha)$ percentile of the $\chi^2(d)$ distribution.

This result is a *large sample approximation*.

The *Wald test*\index{Tests!Wald} is theoretically performed as follows:

1.   Maximize $\log L(\beta_1, \beta_2)$ under ${\cal M}_2$;
     this gives $\log L(\hat{\beta}_1, \hat{\beta}_2)$, and 
     $\hat{\beta}_2$, se($\hat{\beta}_2$).

2.    Calculate the test statistic
      \begin{equation*} 
      T_W = \frac{\hat{\beta}_2}{\mbox{se}(\hat{\beta}_2)}
      \end{equation*}
3.    Under $H_0$, $T_W$ has a *standard normal*
      distribution: $T_W \sim N(0, 1)$.
4.    Reject $H_0$ if the absolute value of $T_W$ is larger than 1.96 on a
      significance level of 5\%.

This is a *large sample approximation*, with the advantage that it is
automatically available in all software. In comparison to the LRT, one
model less has to be fitted. This saves time and efforts, unfortunately on
the expense of accuracy, because it may occasionally give 
*nonsensic  results*. This phenomenon is known as the *Hauck-Donner effect*
[@hd77].\index{Hauck-Donner effect}  $\Box$

### Comparing non-nested models

Non-nested models cannot be compared by a likelihood ratio test, but there are a couple of alternatives 
that are based on comparing maximized likelihood values modified with consideration of the number of 
parameters that needs to be estimated. One such alternative is the *Akaike Information Criterion (AIC)*, see 
@akaike.
