# Survival distributions {#app:B}

The survival distributions we discuss here are all available in **R**. The
most basic survival distribution is the \emph{Exponential distribution},
not because it is useful in demographic applications (it isn't), but
because it is a reference distribution, a special case of other
distributions, and a building block for in many useful models.

The piecewise constant hazard, Gamma, and the Weibull
distributions are 
all generalizations of the exponential.

## Relevant distributions in R

In **R**, there are several families of distributions available. The ones that
are relevant in survival analysis are characterized by having positive
support. For each family of distributions, four functions are available; a
*density* function (name prefix `d`), a *cumulative distribution*
function (name prefix `p`), a *quantile* function (name prefix
`q`), and a *random number* generator (name prefix `r` }).
In the package **eha**, the two functions with prefix `h` and `H` are
added for some distributions. For instance, for the *Weibull*
distribution there are `hweibull` and `Hweibull`, where the first
is the hazard function and the second is the cumulative hazards function.

In the subsection *The Exponential distribution*, we show
exactly what these functions are and how to use them. It should be noted
that in base **R**, there are no survival functions, hazard functions, or
cumulative hazards functions. Some are available in the package **eha**
and others. However, they can easily be derived from the given `p-` and
`d-` functions, as shown in the next subsection.

### The Exponential distribution {#Bexp}
\index{Distributions!exponential|(}
The exponential distribution is characterized by having a constant hazard
rate $lambda$. It has density function
$$
f(t; \lambda) = \lambda e^{-\lambda x}, \quad \lambda > 0, \; t > 0,
$$
and survival function
$$
S(t; \lambda) = e^{-\lambda x}, \quad \lambda > 0, \; t > 0.
$$
The mean is $1/\lambda$ (which, unfortunately, also is the \emph{scale}
parameter\index{scale parameter}), and the variance is $1/\lambda^2$. The exponential 
distribution is representated in **R** by the functions `pexp`,
`dexp`,  `qexp`,  and `rexp`, in order the cumulative distribution
function, the density function, the quantile function (which is the inverse
to the cumulative distribution function), and the random number generator function.

In Figure \@ref(fig:expB) the relevant functions are plotted for an
exponential distribution with $\lambda = 1$ (the default value). It is
created as follows. 

```
library(eha)
x <- seq(0, 4, length = 1000)
oldpar <- par(mfrow = c(2, 2))
plot(x, dexp(x), type = "l", 
     main = "Density", ylab = "")
plot(x, pexp(x, lower.tail = FALSE), type = "l", 
     main = "Survival", ylab = "")
plot(x, hweibull(x, shape = 1), type = "l", 
     main = "Hazard", ylab = "")
plot(x, Hweibull(x, shape = 1), type = "l", 
     main = "Cumulative hazards", ylab = "")
par(oldpar)
```

Note that there are no functions `hexp` or `Hexp`, maybe because
they are too simple, see Figure \@ref(fig:expB). The fact that the
exponential distribution is special case of the Weibull distribution
(`shape = 1`) is utilized.

```{r expB,fig=TRUE,echo=FALSE,height=6,fig.cap="The exponential distribution with scale parameter 1."}
require(eha)
x <- seq(0, 4, length = 1000)
par(mfrow = c(2, 2))
plot(x, dexp(x), type = "l", main = "Density", ylab = "")
plot(x, pexp(x, lower.tail = FALSE), type = "l", main = "Survival", ylab = "")
plot(x, hweibull(x, shape = 1), type = "l", main = "Hazard", ylab = "", ylim = c(0, 1.2))
plot(x, Hweibull(x, shape = 1), type = "l", main = "Cumulative hazards", ylab = "")
```

The exponential distribution is characterized by the fact that it
*lacks memory*. In other words, items whose life lengths follow an
exponential distribution do not age; no matter how old they are, if they
are alive they are as good as new. This concept is not useful when it comes
to human lives, but the life lengths of electronic components are often
modeled by the exponential distribution in reliability
theory\index{reliability theory}.
\index{Distributions!exponential|)}

### The piecewise constant hazard distribution

\index{Distributions!piecewise constant hazard|(}
If the exponential distribution is not useful in describing human lives, it
may be so for short segments of life. At least it will be a good
approximation if the segment is short enough.

This is the idea behind the *piecewise constant hazard*
distribution, called `pch` in **eha**. Its definition involves a
partition of the time (age) axis, and one positive constant (the hazard level)
corresponding to each interval. Note that the last interval will be open,
with infinite length; only a finite number of cutpoints are allowed. The
definition of the hazard function $h$ becomes, with the cuts denoted 
$\mathbf{t} = (t_1, < \cdots < t_n)$ and the levels denoted
$\mathbf{h} = (h_1, \ldots, h_{n+1})$:
\begin{equation}\label{eq:pwcB}
h(t; \mathbf{t}, \mathbf{h}) = \begin{cases}
     h_1 & t \le t_1, \\
     %\cdots & \cdots \\
     h_i & t_{i-1} < t \le t_i, \; i = 2,
     \ldots, n, \\
     %\cdots & \cdots \\
     h_{n+1} & t_n < t. \\
     \end{cases}
\end{equation}
In this definition, the number of levels must be exactly one more than the
number of cutpoints. The relevant functions are shown in
Figure \@ref(fig:pchBout), created as follows:



```{r pchBout,echo=TRUE,height=6,fig.cap="Piecewise constant hazard distribution."}
cuts <- c(1, 2, 3)
n <- length(cuts)
levels <- c(3, 2, 1, 2)
oldpar <- par(mfrow = c(2, 2))
plot(x, dpch(x, cuts = cuts, levels = levels), 
     type = "l", main = "Density", ylab = "", xlab = "t")
plot(x, ppch(x, cuts = cuts, levels = levels, lower.tail = FALSE), 
     type = "l", main = "Survival", ylab = "", xlab = "t")
##plot(x, hpch(x, cuts = cuts, levels = levels),
plot(c(0, cuts[1]), c(levels[1], levels[1]),
     type = "l", main = "Hazard", ylab = "", xlab = "t",
     ylim = c(0, max(levels) + 0.2), xlim = c(0, max(x) + 0.2))
for (i in 1:(n - 1)) lines(c(cuts[i], cuts[i+1]), c(levels[i+1], levels[i+1]))
lines(c(cuts[n], Inf, c(levels[n+1], levels[n+1])))
plot(x, Hpch(x, cuts = cuts, levels = levels), 
     type = "l", main = "Cumulative hazards", ylab = "", xlab = "t")
par(oldpar)
```

 Note that, despite the fact that the hazard function is *not*
 continuous, the other functions are. They are not differentiable at the
 cut points, though.
 
 The piecewise constant hazard distribution is very flexible. It can be made
 arbitrarily close to any continuous distribution by increasing the number
 of cutpoints and choose the levels appropriately. Parametric proportional
 hazards modeling with the *pch* distribution is a serious competitor
 to the Cox regression\index{Cox regression} model, especially with large
 data sets. 
\index{Distributions!piecewise constant hazard|)}

### The Weibull distribution
\index{Distributions!Weibull|(}
The *Weibull* distribution is a very popular parametric model for
survival data, described in detail by Waloddi Weibull [@ww51], but known
earlier. It is one of the so-called extreme-value distributions, and as
such very useful in reliability theory. It is becoming popular in
demographic applications, but in mortality studies it is wise to avoid it
for old age mortality (the hazard grows too slow) and mortality in ages
0-15 years of age (U-shaped hazards, which the Weibull model doesn't allow).

The hazard function of a Weibull distribution is defined by 
$$
h\bigl(t; (p, \lambda)\bigr) = \frac{p}{\lambda}\biggl(\frac{t}{\lambda}\biggr)^{p-1}, \quad t, p,
\lambda > 0.
$$
where $p$ is a *shape* parameter and $\lambda$ is a *scale*
parameter. When $p = 1$, this reduces to $h(t; (1, \lambda) = 1/\lambda$,
which is the exponential distribution with rate $1 / \lambda$. Compare to
the definition of the exponential distribution and note that there
$\lambda$ is the *rate* and here it is a *scale* parameter, which
is the inverted value of the rate.

As mentioned above, the Weibull distribution has a long history in
reliability theory. Early on simple graphical tests were constructed for
judging if a real data set could be adequately described by the Weibull
distribution. The so-called *Weibull paper* was invented.
Starting with the definition of the *cumulative* hazards function $H$,
$$
H\bigl(t; (p, \lambda)\bigr) = \biggl(\frac{t}{\lambda}\biggr)^p,
$$
by taking logarithms of both sides we get 
$$
\ln H\bigl(t; (p, \lambda)\bigr) = p \ln(t) - p \ln(\lambda)
$$
with $x = \ln t$ and $y = \ln H(t)$, this is a straight line. So by
plotting the Nelson-Aalen estimator of the cumulative hazards function
against log time, it is possible to graphically check the Weibull and
exponential assumptions.

```{example, name = "Male mortality and the Weibull distribution"}
```
Is it reasonable to assume that the survival data in the data set
`mort` can be modeled by the Weibull distribution? We construct the
*Weibull plot* with the following code.
```
data(mort)
with(mort, plot(Surv(enter, exit, event), fn = "loglog"))
```
The result is shown in Figure \@ref(fig:wplotB).

```{r wplotB,fig=TRUE,echo=FALSE,fig.cap = "Weibull plot of male mortality data."}
with(mort, plot(Surv(enter, exit, event) ~ 1, fun = "cloglog"),
     xlab = "log(Age - 20)", ylab = "log(Cum. hazards)")
```

Except for the early disturbance, linearity is not too far away (remember
that the log transform magnifies the picture close to zero). Is the slope close to 1? Probably not, the different
scales on the axis makes it hard to judge exactly. We can estimate the
parameters with the \fun{phreg} function.
```{r fitweiB}
fit <- phreg(Surv(enter, exit, event) ~ 1, data = mort)
pr <- rbind(fit$coefficients, sqrt(diag(fit$var)), exp(fit$coefficients))
rownames(pr) <- c("Coef", "se(Coef)", "exp(Coef)")
round(pr, 3)
```
The estimate of the shape parameter is 
`r round(exp(fit$coef[2]), digits = 3)`, and it is significantly
different from 1 ($p = 0.000$),
 so we can firmly reject the hypothesis that
data come from an exponential distribution.
\index{Distributions!Weibull|)}

### The Lognormal distribution
\index{Distributions!lognormal|(}
The lognormal distribution is connected to the normal distribution through
the exponential function: If $X$ is normally distributed, then $Y =
\text{exp}(X)$ is lognormally distributed. Conversely, if $Y$ is
lognormally distributed, then $X = \text{log}(Y)$ is normally distributed.

The lognormal distribution has the interesting property that the hazard
function is first increasing, then decreasing, in contrast to the Weibull
distribution which only allows for monotone (increasing or decreasing)
hazard functions. 

The **R** functions are named `(dpqrhH)lnorm`.
\index{Distributions!lognormal|)}

### The Loglogistic distribution

The loglogistic distribution is very close to the lognormal, but have
heavier tail to the right. Its advantage over the lognormal is that the
hazard function has closed form. It is given by
\begin{equation*}
h(t; (p, \lambda)) = \frac{\frac{p}{\lambda}(\frac{t}{\lambda})^{p-1}}{1 +
  (\frac{t}{\lambda})^p}, \quad t, p, \lambda > 0.
\end{equation*}
With shape $p = 2$ and scale $\lambda = 1$, its appearance is shown in
Figure \@ref(fig:llogB). 

```{r llogB,fig=TRUE,echo=FALSE,fig.cap="Loglogistic hazard function with shape 2 and scale 1."}
x <- seq(0, 10, length = 1000)
plot(x, hllogis(x, shape = 2, scale = 1), 
     type = "l", ylab = "", xlab = "Time")
abline(h = 0)
```

It is produced by the code
```
x <- seq(0, 10, length = 1000)
plot(x, hllogis(x, shape = 2, scale = 1), 
     type = "l", ylab = "", xlab = "Time")
abline(h = 0)
```

The **R** functions are named `(dpqrhH)llogis`.
\index{Distributions!loglogistic|)}

### The Gompertz distribution
\index{Distributions!Gompertz|(}
The Gompertz distribution is useful for modeling old age mortality. The
hazard function is exponentially increasing.
\begin{equation}
h(t; (p, \lambda)) = p \exp\biggl(\frac{t}{\lambda}\biggr), \quad t, p, \lambda > 0.
\end{equation}
It was suggested by @gom25.

The **R** functions are named `(dpqrhH)gamma`.
\index{Distributions!Gompertz|)}

### The Gompertz-Makeham distribution

The Gompertz distribution was generalized by @mak60. The
generalization consists of adding a positive constant to the Gompertz
hazard function,
\begin{equation}
h(t; (\alpha, p, \lambda)) = \alpha + p
\exp\biggl(\frac{t}{\lambda}\biggr), \quad t, \alpha, p, \lambda > 0.
\end{equation}
It is more difficult to work with than the other distributions described
here. It is not one of the possible choices in the functions \fun{phreg} or 
\fun{aftreg} in **eha**.

The **R** functions are named `(dpqrhH)makeham`.
\index{Distributions!Gompertz-Makeham|)}

### The Gamma distribution
\index{Distributions!gamma|(}
The Gamma distribution is another generalization of the exponential
distribution. It is popular in modeling shared 
frailty\index{shared frailty}, see @ph00. It is not one of the
possible distributions 
for the functions \fun{phreg} and \fun{aftreg}, and it is not considered
further.  

The **R** functions are named `(dpqr)gamma`.
\index{Distributions!gamma|)}


## Proportional hazards models {#ph_B}

Proportional hazards families of distributions are characterized by the property
that if a distribution with hazard function $h(t), t > 0$ belongs to the family, so does 
all distributions with hazard functions $\Psi h(t), t > 0; \Psi > 0$. *Natural*
proportional hazards families of distributions include the *Weibull*, *Exponential*, 
*Extreme value*, *Gompertz*, and *Piecewise constant hazard* families of distributions.

As an example, the proportional hazards Gompertz family of distributions is 
characterized by the hazard functions

\begin{equation}
h_{\Psi}(t) = \Psi \exp(\alpha t), \quad t > 0; \; \Psi > 0,
\end{equation}
for each *fixed* $\alpha$, that is, by keeping $\alpha$ fixed and vary 
$\Psi$, one proportional hazards family of distributions is created. We note in 
passing that if $\alpha = 0$, then the resulting family of distributions is *exponential*
with rate $\Psi$, and if $\alpha < 0$, then the distribution is degenerate 
(the probability of "eternal life" is positive). In the `phreg` function in **eha**, 
$\alpha$ is allowed to vary freely, if the parametrization `param = "rate"` is chosen.
Note that it is *not* the default value (it may be in a near future).

Extended, three-parameter, versions of the Lognormal and Loglogistic are fit into 
the proportional hazards framework in the function `phreg. It is simply done by 
adding a multiplicative positive constant to the hazard function as the third parameter.
It is, at the time of this writing, experimental, so use with care.

## Accelerated failure time models {#aft_B}

Accelerated failure time families of distributions are defined by the property that 
if a distribution with survivor function $S(t), t > 0$ belongs to the family, so does
any distribution with survivor function $S(\Psi t), t > 0; \Psi > 0$. Natural families
with this property include the *Lognormal*, *Loglogistic*, *Weibull*, *Extreme value*, 
and *Exponential* families. With a small twist, also the *Gompertz* distribution can be 
(and is, in `aftreg`) included.

