# A Brief Introduction to **R**

```{r settingsC, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(comment = "", message = FALSE, echo = TRUE, cache = FALSE)
options(width = 70, digits = 7)
source("R/fit.out.R")
source("R/tbl.R")
```

There are already many excellent sources of information available for those
who want to learn more about **R**. A visit to the 
[**R** home page](https://www.r-project.org), is highly recommendable. 
When there, look under *Documentation*. The *Task Views* on the home page of 
[CRAN](https://cran.r-project.org) contains a section *Survival* worth examining.

When **R** is installed on a  computer, it is easy to
access the documentation that comes with the distribution; start
**R** and read the help page `?help.start`. A good start for a
beginner is also the book *Introductory Statistics with R* by
[@pd08]. The book on 
*S Programming* [@venrip2] is recommended for the more
advanced studies of the topic that its title implies.

That said, in this chapter we will only look at some aspects of working with **R**,
aspects that are relevant to our topic.

## R in General

In this section we investigate how to work with **R** in general, without
special reference to event history analysis, which will be treated
separately in a later section in this chapter.

### R objects

**R** is an object-oriented programming language, which have certain
implications for terminology. So is for instance everything i
**R** *objects*. We will not draw too much on this fact, but it may be
useful to know.

### Expressions and assignments

Commands are either *expressions*\index{expression} or
*assignments*\index{assignment}. Commands are 
separated by newline (normally) or semi-colon. The hash mark (#) marks the
rest of the line as a *comment*.

```{r ettC}
1 + exp(2.3)
x <- 1 + exp(2.3)
x
```
Note that the **R** *prompt* is ` > `. If the previous expression is
*incomplete*, it changes to ` + `.

The first line above is an expression. It is evaluated as `return` is
pressed, and the result is normally printed. The second line is an
assignment, the result is stored in `x` and not printed. By typing `x`, its
content is normally printed.

The assignment symbol is ` <- `, which consists of two consecutive key
strokes, "<" followed by "-". It is strongly recommended to enclose the
assignment symbol (and any arithmetic operator) between spaces. It is
*forbidden to separate* the two symbols by a space! You will then get
comparison with something negative:
```{r compC}
x <- 3
x < - 3
```
It ia also allowed, nowadays, to use ` = ` as an assignment symbol, but be aware 
of the fact that ` <- ` and ` = ` are not always exchangeable, for instance when 
giving values to arguments in function calls.

### Objects

Everything in **R** is an *object* (object-oriented programming). All
objects have a *mode* and a *length*. Data objects have modes
*numeric*, *complex*, *logical*, or *character*, and
language objects have modes *function*, *call*,
*expression*, *name*, etc. All data objects are *vectors*;
there are *no scalars* in **R**.

Vectors contain elements of the same kind; a *list* can be thought of
a vector where each element can be anything, even a list, and the elements
can be completely different.

There are more properties of objects worth mentioning, but we save that for
the situations where we need them.

### Vectors and matrices

There are five basic types of vectors:
`logical`, `integer`, `double`,
`complex`, and `character`. 
The function `c` (for *concatenate*) creates vectors:
```{r creC}
dat <- c(2.0, 3.4, 5.6)
cols <- c("red", "green", "blue")
```
The elements of a vector may be *named*: 
```{r namC}
names(dat) <- c("Sam", "John", "Kate")
names(dat)
dat
dat["John"]
```
Note the first two lines; `names` may either *display* (second row) or
*replace* (first row).

The function *matrix* creates a matrix (surprise!):
```{r matrC}
Y <- matrix(c(1,2,3,4,5,6), nrow = 2, ncol = 3)
Y
dim(Y) <- c(3, 2)
Y
as.vector(Y)
mode(Y)
attributes(Y)
```

It is also possible to create vectors by the functions `numeric`, 
  `character`, and `logical`. 
```{r ojC}
x <- logical(7)
x
x <- numeric(7)
x
x <- character(7)
x
```
These functions are useful for *allocating* storage (memory), which
later can be filled. This is better practice than letting a vector grow
sequentially. 

The functions `nrow` and `ncol` extracts the number of rows and
columns, respectively, of a matrix.

An *array* is an extension of a matrix to more than two dimensions.

### Lists

Family records may be produced as a list:
```{r famC}
fam <- list(FamilyID = 1, man = "John", wife = "Kate", 
            children = c("Sam", "Bart"))
fam
fam$children
```

### Data frames

A *data frame* is a special case of a *list*. It consists of
variables of the same length, but not necessarily of the same type. Data
frames are created by either `data.frame` or `read.table`, where
the latter reads data from an ASCII file on disk.
```{r dfrC}
dat <- data.frame(name = c("Homer", "Flanders", "Skinner"), 
                  income = c(1000, 23000, 85000))
dat
```
A data frame has *row names* (here: 1, 2, 3) and variable (column)
names (here: `name`, `income`). The data frame is the normal object
for data storage and statistical analysis in **R**.

### Factors

*Categorical* variables are conveniently stored as *factors* in **R**.
```{r facC}
country <- factor(c("se", "no", "uk", "uk", "us"))
country
print.default(country)
```
Factors are internally coded by integers (1, 2, ...). The levels are by
default sorted 
into alphabetical order. Can be changed:
```{r sfacC}
country <- factor(c("se", "no", "uk", "uk", "us"), 
                  levels = c("uk", "us", "se", "no"))
country
```
The first level often has a special meaning ("reference category") in
statistical analyses in **R**.


### Operators

Arithmetic operations are performed on vectors, element by element. The
common operators are ` + - * / ^ `, where ` ^ ` is
exponentiation. For instance
```{r powerC}
x <- c(1, 2, 3)
x^2
y <- 2 * x
y / x
```

### Recycling

The last examples above were examples of *recycling*; If two vectors
of different lengths occur in an expression, the shorter one is recycled
(repeated) as many times as necessary. For instance
```{r recycling}
op <- options(warn = -1) # Turn off warnings.
x <- c(1, 2, 3)
y <- c(2, 3)
x / y
options(op) # Set warnings to original state.
```
so the actual operation performed is
```{r nexC}
c(1, 2, 3) / c(2, 3, 2) # Truncated to equal length
```
It is most common with the shorter vector being of length one, for instance, a
scalar multiplication. If the length of the longer vector is not a multiple
of the length of the shorter vector, a warning is issued: `longer
  object length is not a multiple of shorter object length` (not shown
above). This is often a sign of a mistake in the code (so do not turn off warnings!).

### Precedence

Multiplication and division is performed before addition and subtraction,
as in pure mathematics. The (almost) full set of rules, from highest to
lowest, are:

1.   `$, [[`:  Extraction of elements in a list.
2.   `[`:  Vector extraction.
3.   `^`: Exponentiation. 
4.   `-`: Unary minus.
5.   `:`: Sequence generation.
6.   `%%, %/%, %*%`: Special operators.
7.   `*, /`: & Multiplication and division.
8.   `+, -`: Addition and subtraction.
9.   `<, >, <=, >=, ==, !=`: Comparison.
10.  `!`: Logical negation.
11.  `&, |, &&, ||`: Logical operators.
12.  `$\sim$`: Formula.
13.  `<-`: Assignment.

It is, however, highly recommended to use parentheses often rather than
relying on (and remembering) those rules. Compare, for instance
```{r seqC}
1:3 + 1
1:(3 + 1)
```
For descriptions of specific operators, consult the help system.

## Some Standard **R** Functions

1.   `round`, `floor`, `ceiling`: Rounds to nearest integer,
  downwards, and upwards, respectively.
```{r roundC}
round(2.5) # Round to even
floor(2.5)
ceiling(2.5)
```
2.   `%/%` and `%%` for integer division and modulo reduction.
```{r moduC}
5 %/% 2
5 %% 2
```
3.   Mathematical functions: `abs`, `sign`, `log`, `exp`, etc.
4.   `sum`, `prod`, `cumsum`, `cumprod` for sums and products.
5.   `min`, `max`, `cummin`, `cummax` for extreme values.
6.   `pmin`, `pmax` parallel min and max.
```{r maxC}
x <- c(1, 2, 3)
y <- c(2, 1, 4)
max(x, y)
pmax(x, y)
```
7.   `sort`, `order` for sorting and ordering. Especially useful
  is `order` for rearranging a data frame according to a specific
  ordering of one variable:
```{r orderC}
require(eha)
data(mort)
mt <- mort[order(mort$id, -mort$enter), ]
last <- mt[!duplicated(mt$id), ]
```
First, `mort` is sorted after `id`, and within `id` decreasingly
after `enter` (notice the minus sign in the formula!). Then all rows
with duplicated `id` are removed, only the *first* appearance is
kept. In this way we get a data frame with exactly one row per individual,
the row that corresponds to the individual's last (in age or calendar time).
See also the next item!

8.   `duplicated` and `unique` for marking duplicated elements in
  a vector and removing duplicates, respectively.
```{r duplikatC}
x <- c(1,2,1)
duplicated(x)
unique(x)
x[!duplicated(x)]
``` 


### Sequences

The operator `:` is used to generate sequences of numbers.
```{r colon}
1:5
5:1
-1:5
-(1:5)
-1:-5
``` 
Don't trust that you remember rules of precedence; use parentheses!

There is also the functions `seq` and `rep`, see the help pages!

### Logical expression

*Logical expressions* can take only two distinct values, `TRUE`
and `FALSE` (note uppercase; **R** acknowledges the difference between
lower case and upper case).

Logical vectors may be used in ordinary arithmetic, where they are
*coerced* into numeric vectors with `FALSE` equal to zero and
  `TRUE` equal to one.

### Indexing

*Indexing* is a powerful feature in **R**. It comes in several flavors.


*   **A logical vector** Must be of the same length as the vector it is
  applied to.
```{r vec}
x <- c(-1:4)
x
x[x > 0]
``` 
*   **Positive integers** Selects the corresponding values.
*   **Negative integers** Removes the corresponding values.
*   **Character strings** For named vectors; select the elements with the
  given names.
*   **Empty** Selects all values. Often used to select all rows or columns from a matrix,
```{r indmat}
x <- matrix(c(1,2,3,4), nrow = 2)
x
x[, 1]
x[, 1, drop = FALSE]
``` 

Note that a dimension is lost when one row or column is selected. This can be
overridden by the argument `drop = FALSE` (*don't drop dimension(s)*).

### Vectors and matrices

A *matrix* is a vector with a `dim` attribute.
```{r dim}
x <- 1:4
x
dim(x) <- c(2, 2)
dim(x)
x
t(x)
``` 
Note that matrices are stored *column-wise*. The function `t`
gives the transpose of a matrix.

An example of matrix multiplication.
```{r matmul}
x <- matrix(c(1,2,3,4, 5, 6), nrow = 2)
x
y <- matrix(c(1,2,3,4, 5, 6), ncol = 2)
y
y %*% x
x %*% y
``` 
Dimensions must match according to ordinary mathematical rules.

### Conditional execution

Conditional execution uses the `if` statement (but see also
  `switch`). Conditional constructs are typically used *inside functions*.
```{r if}
x <- 3
if (x > 3) {
    cat("x is larger than 3\n")
}else{ 
    cat("x is smaller than or equal to 3\n")
}
``` 
See also the function `cat` (and `print`).

### Loops

Loops are typically only used *inside functions*.

*   A `for` loop makes an expression to to be iterated as a variable
  assumes all values in a sequence.
```{r for}
x <- 1
for (i in 1:4){
    x <- x + 1
    cat("i = ", i, ", x = ", x, "\n")
}
x
``` 
*   The `while` loop iterates as long as a condition is `TRUE`.
```{r while}
done <- FALSE
i <- 0
while (!done & (i < 5)){
    i <- i + 1
    done <- (i > 3)
}
if (!done) cat("Did not converge\n")
``` 
*   The `repeat` loop iterates indefinitely or until a `break`
  statement is reached.
```{r repeat}
i <- 0
done <- FALSE
repeat{
    i <- i + 1
    done <- (i > 3)
    if (done) break
}
i
``` 


\subsection{Vectorizing}
In **R**, there are a family of functions that operates directly on
vectors. See the documentation of `apply`, `tapply`, `lapply`,
and `sapply`. 


## Writing Functions

Functions are created by assignment using the keyword `function`.
```{r fun}
fun <- function(x) exp(x)
fun(1)
``` 
This function, `fun`, is the exponential function, obviously. Usually,
the function body consists of several statements.
```{r sevC}
fun2 <- function(x, maxit){
    if (any(x <= 0)) error("Only valid for positive x")
    i <- 0
    sum <- 0
    for (i in 0:maxit){
        sum <- sum + x^i / gamma(i + 1)
    }
    sum
}
fun2(1, 5)
fun2(1:5, 5)
```
This function, `fun2`, uses the Taylor expansion of $\exp(x)$ for
calculation. This is a very crude variant, and it is only good for positive
$x$. It throws an error if fed by a negative value. Note that it is
*vectorized*, i.e., the first argument may be a real-valued
vector. Also note the function `any`.

In serious function writing, it is very inconvenient to define the function
at the command prompt. It is much better to write it in a good editor and
then `source` it into the **R** session as needed. If it will be used
extensively in the future, the way to go is to write a `package`. This
is covered later in these notes. Our favorite editor, and the one used in
writing the examples in these notes, is `emacs`, which is free and available on all
platforms. Another option is to use RStudio [@rstudio].

Functions can be defined inside the body of another function. In this way
it is hidden for other functions, which is important when considering
scoping rules. This is a feature that is not available in Fortran
77 or C. Suppose that the function `inside` is defined inside
the function `head`.

1.   A call to `inside` will use the function defined in `head`,
  and not some function outside with the same name.
2.   The function `inside` is not visible outside `head` and
  cannot be used there.

The *return value* of a function is the value of the last expression
in the function body. 

### Calling conventions

The arguments to a function are either *specified* or
*unspecified* (or both). Unspecified arguments are shown as "$\ldots$"
in the function definition. See for instance the functions `c` and `min`.
The "$\ldots$" may be replaced by any number of arguments in a call to the
function.

The *formal* arguments are those specified in the function definition
and the *actual* arguments are those given in the actual call to the
function. The rules by which formal and actual are *matched* are

1.   Any actual argument given as `name = value`, where `name`
  exactly matches a formal argument, is matched.
2.   Arguments specified by `name = value`, but no exact match are
  then considered. If a perfect *partial* match is found, it is accepted.
3.   The remaining unnamed arguments are then matched *in sequence*
  (*positional* matching).
4.   Remaining unmatced actual arguments are part of the ... argument
  if there is one. Otherwise an error occurs.
5.   Formal arguments need not be matched.

  
### The argument `...`

The purpose of the `...` argument is to pass arguments to a
function call inside the actual function. Read *An Introduction to R* (available 
in the **R** help system) for further detail.

### Writing functions \label{sec:fun}

Below is a function that calculates the maximum likelihood estimate of the
parameter in the exponential distribution, given a censored sample.
```
mlexp <- function(x, d){
    n <- length(x)
    ## Check indata:
    if (any(x <= 0)) stop("This function needs positive x")
    if (length(d) != n) stop("Lengths of x and d do not match")
    ## End check

    ## We adopt the rule: d == 0 for censored obs; 1 for 'events'.

    d <- as.numeric(d != 0)

    ## The density function is lambda^d * exp(-lambda * x), x > 0.
    ## But we prefer a parameterization where the 
    ## parameter may vary freely.
    ## Set lambda = exp(alpha):
    ## The density is exp(d * alpha) * exp(-x * exp(alpha)).
    
    sumx <- sum(x)
    sumd <- sum(d)
    loglik <- function(alpha){
        sumd * alpha - sumx * exp(alpha)
    }
    ## We want the first derivative for optimization:
    dloglik <- function(alpha){
        sumd - sumx * exp(alpha)
    }

    ## Start value:
    alpha <- 0
    res <- optim(alpha, fn = loglik, gr = dloglik,
                 method = "BFGS", control = list(fnscale = -1))
    res
}
``` 
First note that the arguments of the function, `x` and `d`, do not
have any default values. Both arguments must be matched in the actual call
to the function. Second, note the check of sanity in the beginning of the
function and the calls to `stop` in case of an error. For less severe
mistakes it is possible to give a `warning`, which does not break the
execution of the function.
```{r runmle, eval = FALSE, echo = FALSE}
exit <- rexp(100, 1)
event <- as.numeric(exit < 2)
exit <- pmax(exit, 2)
res <- mlexp(exit, event)
res
unlist(res)
``` 

### Lazy evaluation

When a function is called, the arguments are parsed, but no evaluated until
it is needed in the function. Specifically, if the function does not use
the argument, it is never evaluated, and the variable need not even exist.
This is in contrast to rules in C and Fortran.
```{r lazy}
fu <- function(x, y) 2 * x
fu(3)
fu(3, y = zz)
exists("zz")
``` 

### Recursion \label{sec:recur}

Functions are allowed to be *recursive*, i.e.,  a function in **R** is allowed
to call itself. See *An Introduction to R* for details, but be warned: This technique
may be extremely memory-intensive and slow. Use with caution, there is often a more
direct way of attack.

### Vectorized functions

Most mathematical functions in **R** are *vectorized*, meaning that if
the argument to the function is a vector, them the result is a vector of
the same length. Some functions apply the recycling rule to vectors shorter
than the longest:
```{r vect}
sqrt(c(9, 4, 25))
pnorm(1, mean = 0, sd = c(1,2,3))
``` 
When writing own functions, the question of making it vectorized should
always be considered. Often it is automatically vectorized, like
```{r vect2}
fun <- function(x) 3 * x - 2
fun(c(1,2,3))
``` 
but not always. Under some circumstances, it may be necessary to
explicitly make the calculations separately for each element in a vector.  

\subsection{Scoping rules}

Within a function in **R**, objects are found in the following order:

1.   It is a \emph{locally} defined variable.
2.   It is in the argument list.
3.   It is defined in the environment where the function is defined.

Note that a function will eventually look for the variable in the work
space. This is often *not* the intention! 

**Tip:** Test functions in a *clean workspace*. For instance, start R with
the flag `R --vanilla`.


## Standard Graphics

The workhorses in standard graphics handling in **R** are the functions 
  `plot` and `par`. Read their respective documentations carefully; you
can manipulate your plots almost without limitation!

See the code and Figure \@ref(fig:plot2C) for the output.
```{r plots}
source("plot2.R")
plot2
``` 


```{r plot2C,fig.cap = "Output from the function plot2."}
plot2()
``` 

## Useful **R** Functions

\subsubsection*{Matching}
\index{Functions!\fun{match}}
There are some useful functions in **R** doing *set operations*. For
exact definitions of these functions, see the documentation.

*   **match**\index{Functions!\fun{match}} `match(x, y)` returns a vector of the same length as 
    `x` with the place of the first occurrence of each element in `x` in
  `y`. If there is mo match for a particular element in `x`,
    `NA` is returned (can be changed). Example
```{r match}
x <- c(1,3,2,5)
y <- c(3, 2, 4, 1, 3)
match(x, y)
```
The `match` function is useful picking information in one data frame
and put it in another (or the same). As an example, look at the data frame
`oldmort`. 
```{r oldmortC}
require(eha)
data(oldmort)
names(oldmort)
``` 
Note the variables `id` (id of current record) and `m.id` (id of
the mother to the current). Some of the mothers are also subjects in this
file, and if we in an analysis want to to use mother's `birthplace` as
a covariate, we can get it from her record (if present; there will be many
missing mothers). This is accomplished with `match`; creating a new
variable `m.birthplace` (mother's birthplace) is as simple as
```{r mbthpC}
oldmort$m.birthplace <- with(oldmort, 
                             birthplace[match(m.id, id)])
``` 
The tricky part here is to get the order of the arguments to `match`
right. Always check the result on a few records!

*   **%in%**\index{Functions!\fun{\%in\%}} Simple version of `match`. Returns a logical vector of
  the same length as `x`.
```{r in}
x %in% y
``` 
A simple and efficient way of selecting subsets of a certain kind, For
instance, to select all cases with parity 1, 2, or 3, use
```{r selC}
data(fert)
f13 <- fert[fert$parity %in% 1:3, ]
``` 
This is the same as
```{r sel2C}
f13 <- fert[(fert$parity >= 1) & (fert$parity <= 3), ]
``` 
Note that the parentheses in the last line are strictly unnecessary, but
their presence increases readability and avoids stupid mistakes about
precedence between operators. Use parentheses often!

*   **subset** Selects subsets of vectors, arrays, and data frames. See the
  help page. 

*   **tapply**  Applies a function to all subsets of a data frame (it is more
  general than that; see the help page). Together with \fun{match} it is
  useful for creating new summary variables for clusters and sticking the
  values to each individual.
  
  If we for all rows in the `fert` data frame want to add the age at
  first birth for the corresponding mother, we can do this as follows.
```{r tapplyC}
data(fert)
indx <- tapply(fert$id, fert$id)
min.age <- tapply(fert$age, fert$id, min)
fert$min.age <- min.age[indx] 
``` 
Check this for yourself! And read the help pages for `match` and
`tapply` carefully.


## Help in **R**

Besides all documentation that can be found on 
[CRAN](https://cran.r-project.org}),
there is an extensive help system online in **R**. To start the general help
system, type

```
> help.start()
```

at the prompt (or, find a corresponding menu item on a graphical user interface system). Then a
help window opens up with access to several FAQs, and all the packages that
are installed.

For immediate help on a specific function you know the name of, say `coxreg`, type
```
> ?coxreg
```
If you only need to see the syntax and input arguments, use the function \fun{args}:
```
> args(coxreg)
``` 

## Functions for Survival Analysis 

Here some useful functions in `eha` and `survival` are listed. In most cases the
description is very sparse, and the reader is recommended to consult the
corresponding help page when more detail is wanted. 

*   **aftreg**\index{Functions!\fun{aftreg}} Fits accelerated failure time
  (AFT) models.
*   **age.window**\index{Functions!\fun{age.window}} Makes a "horizontal
  cut" in the Lexis diagram\index{Lexis disgram}, that is, it selects a
  subset of a data set based on age. The data frame must contain three
  variables that can be input to the
  `Surv`\index{Functions!\fun{Surv}} function. The default names are
  `enter`, `exit`, and `event`. As an example, 
```{r age.windowC}
library(eha)
mort6070 <- age.window(oldmort, c(60, 70))
``` 
limits the study of old age mortality to the (exact) ages between 60 and
70, i.e. from 60.000 up to and including 69.999 years of age.

For *rectangular* cuts in the Lexiz diagram, use both `age.window`
and `cal.window` in succession. The order is unimportant.

*   **cal.window** \index{Functions!\fun{cal.window}} Makes a "vertical
  cut" in the Lexis diagram\index{Lexis disgram}, that is, it selects a
  subset of a data set based on *calendar time*. As an example,
```{r cal.windowC}
library(eha)
mort18601870 <- cal.window(oldmort, c(1860, 1870))
``` 
limits the study of old age mortality to the time period between 1860 and
1870, that is, from January 1, 1860 up to and including December 31, 1869.

For *rectangular* cuts in the Lexiz diagram, use both `age.window`
and `cal.window` in succession. The order is unimportant.

*   **cox.zph**\index{Functions!\fun{cox.zph}} Tests the proportionality assumption on fits from
  `coxph`. This function is part of the survival package.

*   **coxreg**\index{Functions!\fun{coxreg}} Fits Cox proportional hazards
  model. It is a wrapper for `coxph` in case of simpler Cox regression
  models.  

*   **phreg**\index{Functions!\fun{phreg}} Fits parametric proportional
  hazards models.
 
*   **pchreg**\index{Functions!\fun{pchreg}} Piecewise constant proportional hazards 
regression

*   **tpchreg**\index{Functions!\fun{tpchreg}} Piecewise constant proportional hazards 
regression for *tabular data*.

*   **toTpch**\index{Functions!\fun{toTpch}} Tabulating individual survival data. 


*   **risksets**\index{Functions!\fun{risksets}} Finds the members of each
  riskset at all event times.



### Checking the integrity of survival data

In data sources collected through reading and coding old printed
handwritten sources, many opportunities for making and propagating errors occur. Therefore,
it is important to have tools for detecting logical errors like people
dying before they got married, of living at two places at the same time. In
the sources used in this book, such errors occur in almost all data
retrievals. It is important to say that this fact is not a sign of a bad
job in the transfer of data from old sources to digital data files. The
case is almost always that the errors are present in the original source,
and in the registration process no corrections of "obvious" mistakes by
the registrators are allowed; the digital files are supposed to be as close as
possible to the original sources. These are the rules at the Demographic
Database (DDB), Umeå University.

This means that the researcher has a responsibility to check her data for
logical (and other) errors, and truthfully report how she handles these errors.
In my own experience with data from the DDB, the relative frequency of
errors varies between 1 and 5 per cent. In most cases it would not do much
harm to simply delete erroneous records, but often the errors are
"trivial", and it seems easy to guess what the truth should be.

#### Old age mortality

As an example of the kind of errors discussed above, we take a look at the
original version of `oldmort` in `eha`. For survival data in
""interval" form `(enter, exit]`, it is important that enter
is smaller than exit. For individuals with more than one record, the
intervals are not allowed to overlap, and at most one of them is allowed to
end with an `event`. Note that we are talking about survival data,
so there are no repeated events "by design".

The function `check.surv` in `eha` tries to do these kinds of
checks and report back all individuals (`id`) that do not follow the
rules. 
```{r checksC}
load("olm.rda")
require(eha)
errs <- with(olm, check.surv(enter, exit, event, id))
errs
no.err <- length(errs)
``` 
So, `r no.err` individuals have bad records. Let us look at some of them.
```{r lookbadC}
badrecs <- olm[olm$id %in% errs, ]
dim(badrecs)
badrecs[badrecs$id == errs[1], ]
``` 
The error here is that the second interval `(75.526, 87.927]` starts before
the first interval has ended. We also see that this is a married woman who
became a widow in the second record. 

In this case I think that a simple correction saves the situation: Let the
second interval begin where the first ends. The motivation for this choice
is that the first interval apparently ends with the death of the woman's
husband. Death dates are usually reliable, so I believe more in it than in
other dates.

Furthermore, the exact choice here will have very little impact on the
analyses we are interested in. The length of the overlap is less than half
a year, and this person will eventually have the wrong civil status for at
most that amount of time.

It turns out that all "bad" individuals are of this type. Therefore we
"clean" the data set by running it through this function:

```
ger <- function(){
    require(eha)
    load("olm.rda")
    out <- check.surv(olm$enter, olm$exit, olm$event, olm$id)
    ko <- olm[olm$id %in% out, ]
    ## ko is all records belonging to "bad individuals"
    ## Sort ko, by id, then by interval start, then by event
    ko <- ko[order(ko$id, ko$enter, -ko$event), ]
    for (i in ko$id){ # Go thru all 'bad individuals'
        slup <- ko[ko$id == i, , drop = FALSE]
        ## Pick out their records
        n <- NROW(slup)
        if (n > 1){ # Change them
            for (j in (2:n)){
                slup$enter[j] <- slup$exit[j-1]
            }
            ## And put them back:
            ko[ko$id == i, ] <- slup
        }
    }
    ## Finally, put 'all bad' back into place:
    om <- olm
    om[om$id %in% out, ] <- ko
    ## Keep only records that start before they end:
    om <- om[om$enter < om$exit, ]
    om
}
```

Read the comments, lines (or part of lines) starting with \#. The output of
this function, `om`, is the present data file `oldmort` in 
  `eha`, with the errors corrected as indicated above.

Running the function.
```{r rungerC}
source("ger.R")
om <- ger()
left <- with(om, check.surv(enter, exit, event, id))
left
``` 
Still one bad. Let us take a look at that individual.
```{r lookC}
om[om$id == left, ] 
``` 
It seems as if this error (second an third records overlap) should have been
fixed by our function `ger`. However, there may have been a record
removed, so we have to look at the original data file:
```{r origcheckC}
olm[olm$id == left, ]
``` 
Yes, after correction, the third record was removed. The matter can now be fixed
through removing the last record in `om[om\$id == left, ]` :
```{r oncemoreC}
who <- which(om$id == left)
who
om <- om[-who[length(who)], ]
om[om$id == left, ]
``` 
And we are done.


## Reading Data into **R**

The first thing you have to do in a statistical analysis (in **R**) is to get
data into the computing environment. Data may come from different kind of
sources, and **R** has the capacity to read from many kinds of sources, like
data files from other statistical programs, spreadsheet type (think 
*Excel*) data. This is normally done with the aid of the recommended package 
`foreign` [@foreign], but the newer package `haven` [@haven] is also useful.  

If anything else fails, it is almost always possible to get data written in
ASCII format (e.g., *.csv* files). These can be read into **R** with the
function `read.table`.


### Reading data from ASCII files

Ordinary text  (ASCII) files in tabular form can be read into **R** with the
function `read.table`, and the result is a data frame. For instance, with a file looking this:
```{r fuskdataC,echo=FALSE}
xx <- read.table("my.dat", header = TRUE)
print(xx, row.names = FALSE)
``` 
we read it into **R** with
```{r readdataC}
mydata <- read.table("my.dat", header = TRUE)
``` 
Note the argument `header`. It can take two values, `TRUE` or
`FALSE`, with `FALSE` being the default value. If `header =
  TRUE`, then the first row of the data file is interpreted as
*variable names*. 

There are a few optional arguments to `read.table` that are important
to consider. The first is `dec`, which gives the character used as a
decimal point. The default is ".", but in some locales "," (e.g.,
Swedish) is commonly used in output from other software than **R**. The
second is `sep`, which gives the field separator 
character. The default value is "", which means "white space" (tabs,
spaces, newlines, etc.). See the help page for all the possibilities,
including the functions `read.csv`, `read.csv2`, etc., which are
variations of `read.table` with other default values.

The result should always be checked after reading. Do that by printing a
few rows
```{r prrowC}
head(mydata)
``` 
and/or by summarizing the columns
```{r summC}
summary(mydata)
``` 
The `str` function is also helpful:
```{r strC}
str(mydata)
``` 
For each variable, its name and *type* is given, together with the
first few values. In this example it is all the values, and we can also see
that **R** interprets all the values as `integer`. This is fine,
even though at least `enter` and `exit` probably are supposed to be
real-valued. In fact, **R** is not so strict with  numerical types (as
opposed to *C* and *Fortran*). Frequent
*coercing*\index{coercing} takes place, and the
*mode*\index{mode} `numeric` contains
the types `integer` and `double`. 

Now this tends to be confusing, but the bottom line is that you, as a user,
need not worry at all. Except in one place: When you are writing an **R**
function that calls compiled *C* or *Fortran* code, but that
is far beyond the scope of this book. Interested readers are recommended to
read @venrip2.

### Reading foreign data files

Data file from *Minitab*, *SAS*, *SPSS*, *Stata*, etc.,
can be read with a suitable function from the `foreign`
package. Consult its help pages for more information. Another option is to
use one of the `tidyverse` [@tidyverse] packages `haven`, `readxl`, and
`DBI`. For more information, see relevant help pages.

